---
title: "R Notebook"
output: html_notebook
authors: Ria Baipai, Rafaela Drake, Grace McNally
date: "2026-06-11"
---


This analysis file starts from the cleaned analytic dataset created for you:

```{text}
Data/midus_cancer_aging_analysis_dataset.rds
```

Week 1 Tasks:

1. Install R and RStudio
2. Install the following R packages: tidyverse, gtsummary, flextable, broom, officer, mice, corrr, tibble, misty
3. Create ‘Table 1’ univariate statistics on the study population (put this in a Word document)

Installing necessary R packages
# Set Up
```{r}

library(gtsummary)
library(flextable)
library(broom)
library(officer)
library(mice)
library(corrr)
library(tibble)
library(misty)
library(haven)
library(dplyr)
library(readr)
library(tidyverse)
library(labelled)
library(Hmisc)
```


```{r}
theme_gtsummary_journal(journal = "jama")
  # this means all data you create will be saved in the Data folder which is inside this project's working directory

# Use this to get your working directory
project_dir <- getwd()

data_dir  <- file.path(project_dir, "Data")

# this means all outputs you create will be saved in the Ouput folder which is inside this project's working directory
output_dir  <- file.path(project_dir, "Output")
```


The main exposure is standardized perceived cancer risk:

```text
cancerrisk_STD
```

The two outcomes are:

```text
grimage2_STD
dunedinpace_STD
```

To get you started;

# Load analytic dataset

```{r load-data}

library(dplyr) 
# This creates the path where the CSV version will be saved
 # midus_data <- readRDS("~/Desktop/PQAR_Analysis/Data/midus_cancer_aging_analysis_dataset.rds") #Canc_data: data from MIDUS cancer subproject (edited by Ria - added new code for file path)

analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
Canc_data <- readRDS(analysis_path)

glimpse(Canc_data)


  # Setting variables for data table
  age <- Canc_data$age 
  sex <- Canc_data$sex 
  edu <- Canc_data$education 
  income <- Canc_data$income 
  race_ethn <- Canc_data$race_eth 
  mar_stat <- Canc_data$marital
  bmi <- Canc_data$bmi
  grimage2 <- Canc_data$grimage2_STD
  dunedinpace <- Canc_data$dunedinpace_STD
  num_chron <- Canc_data$num_chron
  smoking <- Canc_data$smoking
  alcohol <- Canc_data$alcohol
  famcancer <- Canc_data$famcancer
  cancerrisk <- Canc_data$cancerrisk
  self_efficacy <- Canc_data$efficacy
  
  
  # Create dataframe based on given variables
  df1 <- data.frame(age, race_ethn, sex, mar_stat, edu, income, bmi, grimage2, dunedinpace, num_chron, smoking, alcohol, famcancer, cancerrisk, self_efficacy)
  
# Adding descriptive labels to data frame

  var_label(df1$age) <- "Age"
  var_label(df1$sex) <- "Sex"
  var_label(df1$mar_stat) <- "Marital Status"
  var_label(df1$edu) <- "Education"
  var_label(df1$income) <- "Income"
  var_label(df1$bmi) <- "BMI"
  var_label(df1$race_ethn) <- "Race and Ethnicity"
  var_label(df1$grimage2) <- "GrimAge2"
  var_label(df1$dunedinpace) <- "DunedinPACE"
  var_label(df1$num_chron) <- "Number of Chronic Conditions"
  var_label(df1$smoking) <- "Smoking Status"
  var_label(df1$alcohol) <- "Alcohol Consumption"
  var_label(df1$famcancer) <- "Family Cancer History"
  var_label(df1$self_efficacy) <- "Self Efficacy"
  var_label(df1$cancerrisk) <- "Perceived Cancer Risk"
  colnames(df1) <- Hmisc::label(df1) # Assigning new labels to data frame
  glimpse(df1) 
```



# Create ‘Table 1' univariate statistics on the study population
```{r}

table1 <- midus_data %>% 
  select(race_eth, grimage2, dunedinpace, bmi, age, 
         sex, marital, education, income, num_chron, 
         smoking, alcohol, famcancer, cancerrisk, efficacy) %>%

  tbl_summary( 
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(race_eth, sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ), 
 missing_text = "Missing",
 missing = "ifany",
 digits = all_continuous() ~ 1, 
    label = list(
      race_eth ~ "Race and Ethnicity",
      grimage2 ~ "GrimAge", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "BMI (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education Level",
      income ~ "Annual Income",
      num_chron ~ "Number of Chronic Conditions",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      famcancer ~ "Family History of Cancer",
      cancerrisk ~ "Cancer Risk Score", 
      efficacy ~ "Self-Efficacy Score"
    )
  )  %>%
  #removing missing values for 
  remove_row_type(
    variables = all_continuous(),
    type = "missing"
  ) %>%
  bold_labels() %>%
  modify_caption("**Table I. Univariate Statistics on MIDUS II Participants**")

table1

as_gt(table1) %>% 
 gt::gtsave(filename = "midus_table_1.docx")

```


```{r}
# Create a histogram

histogram_plot <- df1 %>%
  ggplot(aes(x = income)) +
  geom_histogram(bins = 30, color = "white") +
  labs(title = "Distribution of Income",
       x = "Income", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # this centers the title

# Show the plot in the notebook
histogram_plot

# Save the histogram as a PNG image
ggsave(filename = file.path(output_dir, "example_income_histogram.png"),
       plot = histogram_plot, width = 7, height = 5, dpi = 300)
```



# Table 2
```{r}

library(tidyverse)
library(flextable)
# Generate a basic summary table

table2 <- midus_data %>%
  select(race_eth, grimage2, dunedinpace, bmi, age, 
         sex, marital, education, income, num_chron, 
         smoking, alcohol, famcancer, cancerrisk, efficacy) %>%
  tbl_summary(
    by = race_eth,
    type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
     # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    # Define the exact layout styles from your example image
    statistic = list(
      all_continuous() ~  "{mean} ({sd})", # Separate rows for N and Mean
      all_categorical() ~ "{n} ({p}%)"                # Clean fraction format
    ),
    missing_text = "Missing",
    #hiding missing values??
    missing = "ifany", 
    #applying clear labels 
    
    #do we want to use all these variables?? 
    label = list(
      grimage2 ~ "GrimAge", #outcomes 
      dunedinpace ~ "DunedinPACE", #outcomes 
      bmi ~ "BMI (kg/m²)",
      age ~ "Patient Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education Level",
      income ~ "Annual Income ($)",
      num_chron ~ "Number of Chronic Conditions",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      famcancer ~ "Family History of Cancer",
      cancerrisk ~ "Cancer Risk Score", #is this the perceived cancer risk?
      efficacy ~ "Self-Efficacy Score")
    
  )%>%
  bold_labels()
table2
# 2. Convert to gt and apply the precise Word document overrides
as_gt(table2) %>% 
  # Fix 1: Pass caption via gt::tab_caption using a vector to silence Word's counter
  gt::tab_caption(
    caption = c("Table II. Demographic Characteristics of MIDUS Participants by Race and Ethnicity")
  ) %>%

  gt::gtsave(filename = "midus_table_2.docx")

```

# Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)

```{r}

library(tidyverse)
library(gtsummary)
library(gt)


table3 <- midus_data %>%
  select(race_eth, grimage2, dunedinpace, bmi, age, 
         sex, marital, education, income, num_chron, 
         smoking, alcohol, famcancer, cancerrisk, efficacy) %>%
  tbl_summary(
    by = race_eth,
    type = list(
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous2"
    ), #continuous 2 splits into N and Mean (SD) rows 
    # Define the exact layout styles from your example image
    statistic = list(
      all_continuous() ~ c("{N_nonmiss}", "{mean} ({sd})"), # Separate rows for N and Mean
      all_categorical() ~ "{n} ({p}%)"                # Clean fraction format
    ),
    #hiding missing values??
    missing = "ifany", 
    #applying clear labels 
    
    #do we want to use all these variables?? 
    label = list(
      grimage2 ~ "GrimAge", #outcomes 
      dunedinpace ~ "DunedinPACE", #outcomes 
      bmi ~ "BMI (kg/m²)",
      age ~ "Patient Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education Level",
      income ~ "Annual Income ($)",
      num_chron ~ "Number of Chronic Conditions",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      famcancer ~ "Family History of Cancer",
      cancerrisk ~ "Cancer Risk Score", #is this the perceived cancer risk?
      efficacy ~ "Self-Efficacy Score")
    
  )%>%
  add_n() %>% 
  add_p() %>%
  add_stat_label() %>%
  bold_labels() %>%
  as_gt()%>%
  gt::tab_header(
    title = gt::md("**Table 3. Treatment Differences**")) 

table3
```


```{r}
library(corrr)

#Table 3: Zero-Order Correlation Matrix
#Create Table 3: zero-order correlations among age, perceived cancer risk z-score, GrimAge2, DunedinPACE 
#In meeting more variables were discussed* confirm for matrix

#Correlation Matrix -> perceived cancer risk unadjusted relationships between variables with emphasis on outcomes and exposure ()

#using misty package to create correlation matrix
Correlation_Matrix <- cor.matrix(MIDUS_data, age, grimage2, dunedinpace, age, num_chron, cancerrisk, efficacy,
           method = "pearson",
           na.omit = FALSE, sig = FALSE, alpha = 0.05,
           print = c("cor", "p"),
           tri = "lower",
           p.adj = "none",
           digits = 1, p.digits = 3)

#Formatting of row names and convert it to a data frame
table_4_data <- Correlation_Matrix$matrix %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Variable")

#save as word
flextable(table_4_data) %>% 
  autofit() %>%  #formatting
  save_as_docx(path = "Table_4_Correlation_Matrix.docx")
```

```{r}

#Multiple imputation 


#multiple imputation 

library(aplore3)      # ICU dataset
library(tidyverse)    # Data wrangling and visualization
library(mice)         # Multiple imputation
library(missRanger) 


midus_data <- readRDS("~/Desktop/PQAR_Analysis/Data/midus_cancer_aging_analysis_dataset.rds") #Canc_data: data from MIDUS cancer subproject
set.seed(123)
names(midus_data)

# replace w/ confounders 
confounders <- c("grimage2", "dunedinpace", "age", "sex", "race_eth", "marital",
                 "education", "income", "num_chron") 

#missing summary 
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}
missing_summary 

missing_summary(midus_data, confounders)


# Separate the variables we want to impute from the variables we want to keep.
# Here, only the confounders are sent to mice().

to_imp <- midus_data%>% #data cleaning w/ pipe operator 
  select(all_of(confounders))
to_imp #selects the missing values 

keep <- midus_data %>%
  select(-all_of(confounders)) #data w/out cofounders
keep 


imp <- mice( 
  to_imp, #takes in the date
  m = 5, #number of multiple imputations, default 5 
  maxit = 5, #max iterations
  seed = 123, #set a seed to offset random number generator 
  printFlag = TRUE #if true mice will print history
)

# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Combine the non-imputed variables with the imputed confounders
df_mice_imputed <- keep %>%
  bind_cols(df_mice_completed) #binding


missing_summary(df_mice_imputed, confounders)

#View(df_mice_imputed) this is our data frame 
#confirms no missing values - all 0 

############################################################
# 8. Optional: compare original, missing, and imputed missingness
############################################################

#missing_compare <- bind_rows(
#  missing_summary(midus_data, confounders) %>%
#    mutate(dataset = "Original midus data"),
  
#  missing_summary(df_mice_imputed, confounders) %>%
#    mutate(dataset = "Midus after mice imputation"))

#missing_compare %>%
#  select(dataset, variable, percent_missing)

```



