---
title: "R Notebook"
output: html_notebook
authors: Ria Baipai, Rafaela Drake, Grace McNally
date: "2026-06-11"
---


This analysis file starts from the cleaned analytic dataset created for you:

```{text}
Data/midus_cancer_aging_analysis_dataset.rds
```

Week 1 Tasks:

1. Install R and RStudio
2. Install the following R packages: tidyverse, gtsummary, flextable, broom, officer, mice, corrr, tibble, misty
3. Create ‘Table 1’ univariate statistics on the study population (put this in a Word document)


# Set Up

```{r}
# Load needed packages
library(gtsummary)
library(flextable)
library(broom)
library(officer)
library(mice)
library(corrr)
library(tibble)
library(misty)
library(haven)
library(dplyr)
library(readr)
library(tidyverse)
library(labelled)
library(Hmisc)
```


```{r}
# theme for tables
theme_gtsummary_journal(journal = "jama")
```


The main exposure is standardized perceived cancer risk:

```text
cancerrisk_STD
```

The two outcomes are:

```text
grimage2_STD
dunedinpace_STD
```

To get you started;

# Load analytic dataset

```{r load-data}
# Loading the data
midus_data <- readRDS("Data/midus_cancer_aging_analysis_dataset.rds")

# viewing data structure
glimpse(midus_data)

```



# Create ‘Table 1' univariate statistics on the study population

```{r}
# creating overall descriptive statistics
table1 <- midus_data %>% 
  select(age, sex, race_eth, marital, education, income, num_chron, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%

  tbl_summary( 
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(race_eth, sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      race_eth ~ "Race/Ethnicity",
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      num_chron ~ "Number of chronic Conditions",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table I. Univariate Statistics on Study Population")

table1

table1 %>% 
  as_flex_table() %>%
  save_as_docx(path = "Output/Table 1 Overall Descriptives.docx")

```




# Create ‘Table 2' bivariate statistics on the study population
```{r}
# Generate a basic summary table

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, num_chron, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%

tbl_summary(by = race_eth,
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      num_chron ~ "Number of chronic Conditions",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table 2: Descriptive Statistics by Race/Ethnicity on Study Population")

table2

table2 %>% 
  as_flex_table() %>%
  save_as_docx(
    path = "Output/Table 2 Descriptives by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```

# Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)

```{r}

library(tidyverse)
library(gtsummary)
library(gt)


table3 <- midus_data %>%
  select(race_eth, grimage2, dunedinpace, bmi, age, 
         sex, marital, education, income, num_chron, 
         smoking, alcohol, famcancer, cancerrisk, efficacy) %>%
  tbl_summary(
    by = race_eth,
    type = list(
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous2"
    ), #continuous 2 splits into N and Mean (SD) rows 
    # Define the exact layout styles from your example image
    statistic = list(
      all_continuous() ~ c("{N_nonmiss}", "{mean} ({sd})"), # Separate rows for N and Mean
      all_categorical() ~ "{n} ({p}%)"                # Clean fraction format
    ),
    #hiding missing values??
    missing = "ifany", 
    #applying clear labels 
    
    #do we want to use all these variables?? 
    label = list(
      grimage2 ~ "GrimAge", #outcomes 
      dunedinpace ~ "DunedinPACE", #outcomes 
      bmi ~ "BMI (kg/m²)",
      age ~ "Patient Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education Level",
      income ~ "Annual Income ($)",
      num_chron ~ "Number of Chronic Conditions",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      famcancer ~ "Family History of Cancer",
      cancerrisk ~ "Cancer Risk Score", #is this the perceived cancer risk?
      efficacy ~ "Self-Efficacy Score")
    
  )%>%
  add_n() %>% 
  add_p() %>%
  add_stat_label() %>%
  bold_labels() %>%
  as_gt()%>%
  gt::tab_header(
    title = gt::md("**Table 3. Treatment Differences**")) 

table3
```


```{r}
library(corrr)

#Table 3: Zero-Order Correlation Matrix
#Create Table 3: zero-order correlations among age, perceived cancer risk z-score, GrimAge2, DunedinPACE 
#In meeting more variables were discussed* confirm for matrix

#Correlation Matrix -> perceived cancer risk unadjusted relationships between variables with emphasis on outcomes and exposure ()

#using corrr package to create correlation matrix
Correlation_Matrix <- vars %>%
  correlate(method = "pearson", use = "pairwise.complete.obs") %>%
  shave()

# Rename all columns except term to numbers to avoid repition of variable names
names(Correlation_Matrix)[-1] <- as.character(1:6)
print(Correlation_Matrix, digits = 3)

#formatting changes (showing 3 digits for p-values)
Correlation_Matrix_fmt <- Correlation_Matrix %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "", formatC(., format = "f", digits = 3))))

flextable(Correlation_Matrix_fmt) %>%
  autofit() %>%
  save_as_docx(path = "Table_4_Correlation_Matrix.docx")
```

```{r}

#Multiple imputation 

# Loading packages
library(tidyverse)    # Data wrangling and visualization
library(mice)         # Multiple imputation
library(missRanger)   # Random forest imputation
library(VIM)

# Set seed for reproducibility
set.seed(0864)

# Load dataset
data_dir  <- file.path(project_dir, "Data")
output_dir  <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
library(dplyr)

# Compiling relevant confounders in vector
# i.e. subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates
midus_vars <- midus_data[ , c("grimage2_STD", "dunedinpace_STD", "cancerrisk_STD",
                 "race_eth", "age", "sex", "education", "income",
                 "smoking", "bmi", "marital", "famcancer")]

confounders <- confounders <- c("age", "sex", "education", "income",
  "smoking", "bmi", "famcancer", "marital")

# Summarizing missing data
md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}
# missing_summary 

missing_summary(midus_data, confounders)

# Generating predictor matrix
quickpred(midus_vars)

# Creating a methods object
midus_methods <- make.method(midus_vars[confounders]) 
midus_methods

# Assigning new methods to categorical variables
midus_methods["education"] <- "polyreg"
midus_methods["famcancer"] <- "logreg"
midus_methods["marital"] <- "polyreg"

# Separate the variables we want to impute from the variables we want to keep.
# Here, only the confounders are sent to mice().

to_imp <- midus_data%>% #data cleaning w/ pipe operator 
  select(all_of(confounders))
to_imp #selects the missing values 

keep <- midus_data %>%
  select(-all_of(confounders)) #data w/out cofounders
keep 


imp <- mice(
  to_imp, # data: what we want to impute
  m = 20, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Combine the non-imputed variables with the imputed confounders
df_mice_imputed <- keep %>%
  bind_cols(df_mice_completed) #binding


missing_summary(df_mice_imputed, confounders)

#View(df_mice_imputed) this is our data frame 
#confirms no missing values - all 0 

############################################################
# 8. Optional: compare original, missing, and imputed missingness
############################################################

#missing_compare <- bind_rows(
#  missing_summary(midus_data, confounders) %>%
#    mutate(dataset = "Original midus data"),
  
#  missing_summary(df_mice_imputed, confounders) %>%
#    mutate(dataset = "Midus after mice imputation"))

#missing_compare %>%
#  select(dataset, variable, percent_missing)

```



