---
title: "R Notebook"
output: html_notebook
authors: Ria Baipai, Rafaela Drake, Grace McNally
date: "2026-06-11"
---


This analysis file starts from the cleaned analytic dataset created for you:

```{text}
Data/midus_cancer_aging_analysis_dataset.rds
```

Week 1 Tasks:

1. Install R and RStudio
2. Install the following R packages: tidyverse, gtsummary, flextable, broom, officer, mice, corrr, tibble, misty
3. Create ‘Table 1’ univariate statistics on the study population (put this in a Word document)


# Set Up

```{r}
# Load needed packages

library(gtsummary)
library(flextable)
library(broom)
library(officer)
library(mice)
library(corrr)
library(tibble)
library(misty)
library(haven)
library(dplyr)
library(readr)
library(tidyverse)
library(labelled)
library(Hmisc)
library(modelsummary)
library(naniar) #mcar test
# theme for tables
library(ggeffects) #for making plots
library(ggplot2)
library(patchwork)
library(patchwork) #for ordering figures in rows and columns
theme_gtsummary_journal(journal = "jama")
```


```{r}
# theme for tables
theme_gtsummary_journal(journal = "jama")
```


The main exposure is standardized perceived cancer risk:

```text
cancerrisk_STD
```

The two outcomes are:

```text
grimage2_STD
dunedinpace_STD
```

To get you started;

# Load analytic dataset

```{r load-data}
# Loading the data
midus_data <- readRDS("Data/midus_cancer_aging_analysis_dataset.rds")

# viewing data structure
glimpse(midus_data)

```








# Create ‘Table 1' univariate statistics on the study population

```{r}
# creating overall descriptive statistics
table1 <- midus_data %>% 
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
  tbl_summary( 
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(race_eth, sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p})" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss})",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = list(all_continuous() ~ c(1,1), 
                  all_categorical() ~ c(0,1)
    ),
    label = list(
      race_eth ~ "Race/Ethnicity",
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table I. Univariate Statistics on Study Population")

table1

table1 %>% 
  as_flex_table() %>%
  save_as_docx(path = "Output/Grace_Table 1 Overall Descriptives.docx")

```
#Table 2 

# Create ‘Table 2' bivariate statistics on the study population
```{r}
# Generate a basic summary table

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
tbl_summary(by = race_eth,
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss})",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table 2: Descriptive Statistics by Race/Ethnicity on Study Population")

table2

table2 %>% 
  as_flex_table() %>%
  save_as_docx(
    path = "Output/Grace_Table 2 Descriptives by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```


# Table 3: 

```{r}

library(tidyverse)
library(gtsummary)
library(gt)



table3 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
tbl_summary(by = race_eth,
   type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical"),

    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p})" 
    ),

    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 

    #hiding missing values??
    #applying clear labels 
    
    #do we want to use all these variables?? 
   label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>% 
  add_overall() %>% 
      add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list( 
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  add_stat_label() %>%
  bold_labels() %>%
  modify_caption("Table 3: Descriptive Statistics Overall and Stratified by Race and Ethnicity")

table3

table3 %>% 
  as_flex_table() %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  save_as_docx(
    path = "Output/Grace_Table 3 Descriptive Overall and by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```


#Table 4: correlation matrix

```{r}
#Table 4: Zero-Order Correlation Matrix -> perceived cancer risk unadjusted relationships between variables with #emphasis on outcomes and exposure
cor_vars <- midus_data %>%
  select(grimage2, dunedinpace, age, cancerrisk_STD, efficacy) %>%
 filter(  !is.na(grimage2) #dropping missing value from grimage2
  )
#Convert to a matrix
vars_matrix <- as.matrix(cor_vars)

#Correlation matrix using Hmisc
correlation_results <- Hmisc::rcorr(vars_matrix, type=c("pearson"))
#extract correlation matrix
r_matrix <- correlation_results$r
#extract p-values
p_matrix <- correlation_results$P

#adding starts for p-values and keep lower triagnle
r_formatted <- format(round(r_matrix, 3), nsmall = 3) #issue
stars <- ifelse(p_matrix <0.001, "***",
                ifelse(p_matrix < 0.01, "**",
                       ifelse(p_matrix <0.05, "*", "")))
#combined
final_cor_matrix <- matrix(paste0(r_formatted, stars), nrow = nrow(r_matrix))
rownames(final_cor_matrix) <- rownames(r_matrix)
colnames(final_cor_matrix) <- colnames(r_matrix)

#remove self-correlations
diag(final_cor_matrix) <- ""

#only keep lower triangle
final_cor_matrix[upper.tri(final_cor_matrix)] <- ""

#convert to df and format row names as a separate column
final_df <- as.data.frame(final_cor_matrix)
final_df <- tibble::rownames_to_column(final_df, var = "Variable")

#Professional variable labels
#vector mapping old names to new labels
clean_labels <- c(
  "grimage2"        = "GrimAge2 z-score",
  "dunedinpace"     = "DunedinPACE z-score",
  "age"             = "Age",
  "cancerrisk_STD"  = "Perceived cancer risk z-score",
  "efficacy"        = "Self-efficacy for Cancer Prevention"
)

#rename in final df
final_df$Variable <- clean_labels[final_df$Variable]
#add numbers to row names
final_df$Variable <- paste0(1:nrow(final_df), ". ", final_df$Variable) 
#change column headers to numbers (corresponding with row numbering)
colnames(final_df)[2:ncol(final_df)] <- 1:(ncol(final_df) - 1)


#formatting (flextable)
ft <- flextable(final_df) %>%
  #set header labels
  set_header_labels(
    Variable = "Variable",
    grimage2 = "1",
    dunedinpace = "2",
    age = "3",
    num_chron = "4",
    cancerrisk_STD = "5",
    efficacy = "6") %>%
  theme_booktabs() %>% #APA
  bold(part = "header") %>%
  autofit() %>%
  fit_to_width(max_width = 6.5, unit = "in") %>%
  align(j = 1, align = "left", part = "all") %>% 
  align(j = 2:ncol(final_df), align = "center", part = "all") %>%
  # Add a table title
  add_header_lines("1.1. Table 3. Zero-order correlations among epigenetic aging measures, perceived cancer risk, age, self-efficacy, and chronic conditions.") %>%
  #footer with stars info
  add_footer_lines("* p < 0.05, ** p < 0.01, *** p < 0.001") %>%
  font(fontname = "Arial", part = "all")

ft

#export
save_as_docx(ft, path = "Output/Grace_Table_4_Correlation_Matrix.docx")
```



# Multiple Imputation


```{r}
# Conducting multiple imputation 

# Subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates
#15 vars total 
# checking for row containing NA value in grimAge2_STD
print(is.na(midus_data$grimage2_STD))

midus_vars <- midus_data %>% 
              select("income", "famcancer", "efficacy",
                     "marital", "education", "grimage2_STD", "dunedinpace_STD",
                     "cancerrisk_STD", "race_eth", "age", "sex",
                     "smoking", "alcohol", "bmi") %>%
  # Removing row with NA value for grimage2_STD
  filter(
    !is.na(grimage2_STD) #dropping missing value from grimage2
  ) %>%
  # Categorical variables should be coded as factors in order to invoke polyreg
  mutate(
      race_eth = factor(race_eth),
      alcohol = factor(alcohol),
      sex = factor(sex),
      education = factor(education, ordered = TRUE),
      smoking = factor(smoking),
      marital = factor(marital),
      famcancer = factor(famcancer)
    )


# confirming NA value has been removed
print(is.na(midus_vars$grimage2_STD))

# Run Little's MCAR
library(naniar)  # you need this package to run MCAR (Roph)
mcar_test(midus_vars)
# statistic df  p-value  missing.patterns
# 212.3461	162	0.004844206	13	

# Create a clean character vector tracking all variables in our target analysis subset
analysis_variables <- names(midus_vars)

md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns 
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

#missing summary function 
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    ) %>%
  arrange(desc(percent_missing)) #arrange output in descending over 
}


missing_summary(midus_data, analysis_variables)

#predictor matrix 
pred <- mice(midus_vars, maxit = 0)$predictorMatrix 

# Creating a methods object
midus_methods <- make.method(midus_vars) 
midus_methods # Looking to check method types for categorical vars

# Assigning new methods to categorical variables
midus_methods["education"] <- "polr" # changed this from "polyreg" as education is an ordered categorical variable
#polyreg used for categorical values with 2+ categories
midus_methods["famcancer"] <- "logreg" #logreg for nominal categorical values
midus_methods["marital"] <- "polyreg"
midus_methods # Checking to ensure variables are updated

imp <- mice(
  midus_vars, # data: what we want to impute
  m = 10, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed


# Re-run missing summary on the complete dataset to verify all missingness dropped to 0%
missing_summary(df_mice_completed, analysis_variables)

```
#regression analysis with interaction term in all models
```{r}

# =========================================================================
# 1. REGRESSION MODELS (Overall Sample)
# =========================================================================
df_overall <- df_mice_completed

# --- Outcome: GrimAge2 ---
grim_ov_m1 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital, data = df_overall)
grim_ov_m2 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income, data = df_overall)
grim_ov_m3 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer, data = df_overall)
grim_ov_m4 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy, data = df_overall)
grim_ov_m5 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)


# --- Outcome: DunedinPACE ---
pace_ov_m1 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital, data = df_overall)
pace_ov_m2 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income, data = df_overall)
pace_ov_m3 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer, data = df_overall)
pace_ov_m4 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy, data = df_overall)
pace_ov_m5 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)


# Create two flat lists containing all 6 models
grim_list <- list("M1"=grim_ov_m1, "M2"=grim_ov_m2, "M3"=grim_ov_m3, "M4"=grim_ov_m4, "M5"=grim_ov_m5)
pace_list <- list("M1"=pace_ov_m1, "M2"=pace_ov_m2, "M3"=pace_ov_m3, "M4"=pace_ov_m4, "M5"=pace_ov_m5)

# Variable mappings
# Variable mappings (Updated to handle both R naming conventions)
variables_to_show <- c(
  # Main Effect
  "cancerrisk_STD"                              = "Perceived Cancer Risk (Main Effect)",
  
  # Interaction terms (Continuous : Factor order)
  "cancerrisk_STD:race_ethnon-Hispanic Black"   = "Perceived cancer risk * non-Hispanic Black",
  "cancerrisk_STD:race_ethHispanic"             = "Perceived cancer risk * Hispanic",
  "cancerrisk_STD:race_ethnon-Hispanic Other"   = "Perceived cancer risk * non-Hispanic Other",
  
  # Interaction terms (Factor : Continuous order backup)
  "race_ethnon-Hispanic Black:cancerrisk_STD"   = "Perceived cancer risk * non-Hispanic Black",
  "race_ethHispanic:cancerrisk_STD"             = "Perceived cancer risk * Hispanic",
  "race_ethnon-Hispanic Other:cancerrisk_STD"   = "Perceived cancer risk * non-Hispanic Other"
)

# =========================================================================
# 2. EXTRACT NATIVE DATA FRAMES FROM MODELSUMMARY
# =========================================================================
df_grim <- modelsummary(
  grim_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}\n[{conf.low}, {conf.high}]", statistic = NULL,
  coef_map = variables_to_show
) %>%
  mutate(`Outcome Metric` = "GrimAge2")

df_pace <- modelsummary(
  pace_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}\n[{conf.low}, {conf.high}]", statistic = NULL,
  coef_map = variables_to_show
) %>%
  mutate(`Outcome Metric` = "DunedinPACE")

# =========================================================================
# 3. MERGE AND CLEAN ROW DATA SHAPES
# =========================================================================
final_df <- bind_rows(df_grim, df_pace) %>%
  # Keep columns M1 through M6 intact
  select(`Outcome Metric`, Predictor = term, M1, M2, M3, M4, M5)

# =========================================================================
# 4. BUILD THE FINAL SEAMLESS FLEXTABLE
# =========================================================================
final_table <- flextable(final_df) %>%
  merge_v(j = "Outcome Metric") %>% 
  valign(j = "Outcome Metric", valign = "top") %>%
  add_footer_lines("+ p < 0.10, * p < 0.05, ** p < 0.01, *** p < 0.001.") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  set_caption(
    caption = "Table 4: Overall models with standardized beta coefficients (and 95% confidence intervals) from a series of linear regression models quantifying the association between perceived cancer risk and epigenetic aging among MIDUS II sample"
  )

# =========================================================================
# 5. EXPORT TO WORD
# =========================================================================
if(!dir.exists("Output")) dir.create("Output")

save_as_docx(
  final_table,
  path = "Output/grace_overall_models.docx",
  pr_section = prop_section(
    page_size = page_size(orient = "landscape", width = 11.7, height = 8.3)
  )
)

```

#regression analysis with interaction term only model 6

```{r}
# =========================================================================
# 1. REGRESSION MODELS (Overall Sample)
# =========================================================================
df_overall <- df_mice_completed

# --- Outcome: GrimAge2 ---
grim_ov_m1 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital, data = df_overall)
grim_ov_m2 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income, data = df_overall)
grim_ov_m3 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer, data = df_overall)
grim_ov_m4 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer + efficacy, data = df_overall)
grim_ov_m5 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)
grim_ov_m6 <- lm(grimage2_STD ~ age + sex + race_eth + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi + (cancerrisk_STD * race_eth), data = df_overall)

# --- Outcome: DunedinPACE ---
pace_ov_m1 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital, data = df_overall)
pace_ov_m2 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income, data = df_overall)
pace_ov_m3 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer, data = df_overall)
pace_ov_m4 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer + efficacy, data = df_overall)
pace_ov_m5 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)
pace_ov_m6 <- lm(dunedinpace_STD ~ age + sex + race_eth + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi + (cancerrisk_STD * race_eth), data = df_overall)

# Create two flat lists
grim_list <- list("M1"=grim_ov_m1, "M2"=grim_ov_m2, "M3"=grim_ov_m3, "M4"=grim_ov_m4, "M5"=grim_ov_m5, "M6"=grim_ov_m6)
pace_list <- list("M1"=pace_ov_m1, "M2"=pace_ov_m2, "M3"=pace_ov_m3, "M4"=pace_ov_m4, "M5"=pace_ov_m5, "M6"=pace_ov_m6)

# Variable mappings
variables_to_show <- c(
  "cancerrisk_STD"                            = "Perceived Cancer Risk (Main Effect)",
  "race_ethnon-Hispanic Black:cancerrisk_STD" = "Perceived cancer risk * non-Hispanic Black",
  "race_ethHispanic:cancerrisk_STD"           = "Perceived cancer risk * Hispanic",
  "race_ethnon-Hispanic Other:cancerrisk_STD" = "Perceived cancer risk * non-Hispanic Other"
)

# =========================================================================
# 2. EXTRACT NATIVE DATA FRAMES FROM MODELSUMMARY
# =========================================================================
# By using output = "data.frame", we escape the list processing errors entirely
df_grim <- modelsummary(
  grim_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}\n[{conf.low}, {conf.high}]", statistic = NULL,
  coef_map = variables_to_show
) %>%
  mutate(`Outcome Metric` = "GrimAge2")

df_pace <- modelsummary(
  pace_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}\n[{conf.low}, {conf.high}]", statistic = NULL,
  coef_map = variables_to_show
) %>%
  mutate(`Outcome Metric` = "DunedinPACE")

# =========================================================================
# 3. MERGE AND CLEAN ROW DATA SHAPES
# =========================================================================
final_df <- bind_rows(df_grim, df_pace) %>%
  # Position "Outcome Metric" first on the left, followed by the Predictor (term)
  select(`Outcome Metric`, Predictor = term, M1, M2, M3, M4, M5, M6)
  

# =========================================================================
# 4. BUILD THE FINAL SEAMLESS FLEXTABLE
# =========================================================================
final_table <- flextable(final_df) %>%

  # Merge duplicate Outcome Metric labels vertically for a clean visual block
  merge_v(j = "Outcome Metric") %>% 
  valign(j = "Outcome Metric", valign = "top") %>%
  # Match your document style guidelines
  add_footer_lines("* p < 0.05, ** p < 0.01, *** p < 0.001.") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  set_caption(
    caption = "Table 4: Overall models with standardized beta coefficients (and 95% confidence intervals) from a series of linear regression models quantifying the association between perceived cancer risk and epigenetic aging among MIDUS II sample"
  )

# =========================================================================
# 5. EXPORT TO WORD
# =========================================================================
if(!dir.exists("Output")) dir.create("Output")

save_as_docx(
  final_table,
  path = "Output/grace_overall_models2.docx",
  pr_section = prop_section(
    page_size = page_size(orient = "landscape", width = 11.7, height = 8.3)
  )
)

```

#overall/supplemental table Dunedin PACE
```{r}
#baseline summary simplest model 
modelsummary(grim_list)
modelsummary(pace_list)

#creating publication ready tables 

#DunedinPACE data 

#coefficient map 
# Map your R coefficients to the exact display names from the table
coef_map <- c(
  "cancerrisk_STD" = "Perceived cancer risk (z-score)",
  "age" = "Age",
  "sexFemale" = "Female",
  "race_ethnon-Hispanic Black" = "non-Hispanic Black",
  "race_ethHispanic" = "Hispanic",
  "race_ethnon-Hispanic Other" = "non-Hispanic Other",
  "maritalDivorced/Separated/Widowed" = "Divorced/Separated/Widowed",
  "maritalNever married" = "Never married",
  "educationsome college/associate's degree" = "Some college/associate's degree",
  "educationcollege degree or higher" = "College degree or higher",
  "income" = "Annual household income",
  "famcancerYes" = "Yes",
  "efficacy" = "Self-efficacy for cancer prevention",
  "smokingPast" = "Past",
  "smokingCurrent" = "Current",
  "alcohol<1 day a week" = "<1 day a week",
  "alcohol1-2 days a week" = "1-2 days a week",
  "alcohol3+ days a week" = "3+ days a week",
  "bmi" = "Body mass index (kg/m²)",
  "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (z-score) * non-Hispanic Black",
  "cancerrisk_STD:race_ethHispanic" = "Perceived cancer risk (z-score) * Hispanic",
  "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (z-score) * non-Hispanic Other"
)

# 1. Generate the base table with modelsummary
raw_table <- modelsummary(
  pace_list,
  output = "gt",
  estimate = "{estimate} ({conf.low} to {conf.high})", # Formats as Beta (95% CI)
  statistic = NULL,                                    # Removes standard errors row
  fmt = 2,
  coef_map = coef_map,
  gof_map = NA,                                        # Removes R2, AIC, etc. from bottom
  title = "Supplemental Table. Adjusted associations between perceived cancer risk and DunedinPACE among 1,155 participants from the Midlife in the United States Study."
)

# 2. Corrected gt code (removes the broken tab_cols_merge function)
final_table_dune <- raw_table %>%
  cols_label(
    "M1" = html("Model 1<br><small>Beta (95% CI)</small>"),
    "M2" = html("Model 2<br><small>Beta (95% CI)</small>"),
    "M3" = html("Model 3<br><small>Beta (95% CI)</small>"),
    "M4" = html("Model 4<br><small>Beta (95% CI)</small>"),
    "M5" = html("Model 5<br><small>Beta (95% CI)</small>"),
    "M6" = html("Model 6<br><small>Beta (95% CI)</small>")
  ) %>%
  # Align text left for labels, center for numbers
  cols_align(align = "left", columns = 1) %>%
  cols_align(align = "center", columns = 2:7) %>%
  # Add footnotes/abbreviations at the bottom
  tab_source_note(source_note = "Abbreviation: CI = Confidence Interval") %>%
  tab_source_note(source_note = "Abbreviation: CI = Confidence interval. Results were pooled across ten imputed datasets.") %>%
  # Apply clean APA/Journal style styling
  tab_options(
    table.border.top.color = "black",
    table.border.top.width = px(2),
    table.border.bottom.color = "black",
    table.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(1.5),
    table_body.border.bottom.color = "black",
    table_body.border.bottom.width = px(1.5),
    heading.title.font.weight = "bold",
    heading.title.font.size = pct(110)
  )

# Display the fixed table
final_table_dune

# =========================================================================
# EXPORT GT TABLE DIRECTLY TO WORD 
# =========================================================================
# Ensure the output folder exists before exporting
if(!dir.exists("Output")) dir.create("Output")

# Save the gt table directly as a .docx file
gtsave(
  data = final_table_dune,
  filename = "grace_all_associations_dunedinpace.docx",
  path = "Output"
)


```


#overall/supplemental table for grimage2

```{r}

# =========================================================================
# PUBLICATION READY SUPPLEMENTAL TABLE: GrimAge2
# =========================================================================

# Complete Coefficient Map for all explicit dummy levels generated by R
coef_map <- c(
  "cancerrisk_STD"                            = "Perceived cancer risk (z-score)",
  "age"                                       = "Age",
  "sexFemale"                                 = "Female",
  "race_ethnon-Hispanic Black"                = "non-Hispanic Black",
  "race_ethHispanic"                          = "Hispanic",
  "race_ethnon-Hispanic Other"                = "non-Hispanic Other",
  "maritalDivorced/Separated/Widowed"         = "Divorced/Separated/Widowed",
  "maritalNever married"                      = "Never married",
  "educationsome college/associate's degree"  = "Some college/associate's degree",
  "educationcollege degree or higher"         = "College degree or higher",
  "income"                                    = "Annual household income",
  "famcancerYes"                              = "Yes",
  "efficacy"                                  = "Self-efficacy for cancer prevention",
  "smokingPast"                               = "Past",
  "smokingCurrent"                            = "Current",
  "alcohol<1 day a week"                      = "<1 day a week",
  "alcohol1-2 days a week"                    = "1-2 days a week",
  "alcohol3+ days a week"                     = "3+ days a week",
  "bmi"                                       = "Body mass index (kg/m²)",
  "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (z-score) * non-Hispanic Black",
  "cancerrisk_STD:race_ethHispanic"           = "Perceived cancer risk (z-score) * Hispanic",
  "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (z-score) * non-Hispanic Other"
)

# 1. Generate the base table with modelsummary for GrimAge2
raw_table_grim <- modelsummary(
  grim_list,
  output = "gt",
  estimate = "{estimate} ({conf.low} to {conf.high})", # Formats as Beta (95% CI)
  statistic = NULL,                                    # Removes standard errors row
  fmt = 2,
  coef_map = coef_map,
  gof_map = NA,                                        # Removes R2, AIC, etc. from bottom
  title = "Supplemental Table. Adjusted associations between perceived cancer risk and GrimAge2 among 1,155 participants from the Midlife in the United States Study."
)

# 2. Format the gt table object for professional publication
final_table_grim <- raw_table_grim %>%
  cols_label(
    "M1" = html("Model 1<br><small>Beta (95% CI)</small>"),
    "M2" = html("Model 2<br><small>Beta (95% CI)</small>"),
    "M3" = html("Model 3<br><small>Beta (95% CI)</small>"),
    "M4" = html("Model 4<br><small>Beta (95% CI)</small>"),
    "M5" = html("Model 5<br><small>Beta (95% CI)</small>"),
    "M6" = html("Model 6<br><small>Beta (95% CI)</small>")
  ) %>%
  # Align text left for labels, center for numbers
  cols_align(align = "left", columns = 1) %>%
  cols_align(align = "center", columns = 2:7) %>%
  # Add footnotes/abbreviations at the bottom
  tab_source_note(source_note = "Abbreviation: CI = Confidence interval. Results were pooled across ten imputed datasets.") %>%
  # Apply clean APA/Journal style styling
  tab_options(
    table.border.top.color = "black",
    table.border.top.width = px(2),
    table.border.bottom.color = "black",
    table.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(1.5),
    table_body.border.bottom.color = "black",
    table_body.border.bottom.width = px(1.5),
    heading.title.font.weight = "bold",
    heading.title.font.size = pct(110)
  )

# Display the fixed GrimAge2 table
final_table_grim <- gt() %>%
  tab_header(
    title = "Motor Trend Car Road Tests",

# =========================================================================
# EXPORT GT TABLE DIRECTLY TO WORD 
# =========================================================================
# Ensure the output folder exists before exporting
if(!dir.exists("Adjusted associations between perceived cancer risk and GrimAge2  ")) dir.create("Output")

# Save the gt table directly as a .docx file
gtsave(
  data = final_table_grim,
  filename = "grace_all_associations_grimage.docx",
  path = "Output"
)

```



# model 4 figure
```{r} 
#install.packages("ggeffects")
#install.packages("patchwork") 

# =========================================================================
# 1. GENERATE MARGINAL PREDICTIONS FOR MODEL 4
# =========================================================================
# Re-extracting from M4 directly inside your named list objects

pred_grim <- predict_response(
  grim_list$M4, 
  terms = c("cancerrisk_STD [-2,-1,0,1,2]", "race_eth")
)

pred_pace <- predict_response(
  pace_list$M4, 
  terms = c("cancerrisk_STD [-2,-1,0,1,2]", "race_eth")
)

# Convert to data frames for custom ggplot handling
df_plot_grim <- as.data.frame(pred_grim)
df_plot_pace <- as.data.frame(pred_pace)

# Define matching professional palette from your example figure
color_palette <- c(
  "non-Hispanic Black" = "#377eb8",  # Clean Blue
  "non-Hispanic Other" = "#ff7f00",  # Sharp Orange
  "Hispanic"           = "#4daf4a",  # Muted Green
  "non-Hispanic White" = "#e41a1c"   # Red
)

# =========================================================================
# 2. PLOT A: GrimAge2 (Model 4 predictions)
# =========================================================================
plot_a <- ggplot(df_plot_grim, aes(x = x, y = predicted, color = group, group = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = color_palette) +
  scale_x_continuous(breaks = -2:2, labels = c("-2", "-1", "0", "1", "2")) + #breaks adds ticks 
  #scale_y_continuous(limits = c(-1.8, 1.8), breaks = seq(-1.5, 1.5, 0.5), labels = c("-1.5", "1.0","0.5", "0.0", "+0.5", "+1.0", "+1.5")) +
  labs(
    title = "(A)",
    x = "Perceived Cancer Risk (z-score)",
    y = "Predicted GrimAge2 (z-score)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "none" # Hidden on left plot; managed globally by patchwork
  )

#dunedin pace plot for model 4 
plot_b <- ggplot(df_plot_pace, aes(
  x = x, 
  y = predicted, 
  color = group, 
  group = group
)) + geom_line(linewidth = 1.2) + 
  geom_point(size = 3) + 
  scale_color_manual(values = color_palette) + 
  scale_x_continuous(breaks = -2:2, labels = c("-2", "-1", "0", "1", "2")) + #breaks adds ticks 
  #scale_y_continuous(limits = c(-1.8, 1.8), breaks = seq(-1.5, 1.5, 0.5), labels = c("-1.5", "-1.0", "-0.5", "0.0", "+0.5", "+1.0", "+1.5")) +
  labs(
    title = "(B)",
    x = "Perceived Cancer Risk (z-score)",
    y = "Predicted DunedinPACE (z-score)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "right" 
    # Hidden on left plot; managed globally by patchwork
  )


wrapped_title <- str_wrap(
  "Association Between Perceived Cancer Risk and Biological Aging Metrics with Regression Model 4", 
  width = 60 # Restricts lines to ~60 characters max
)

# 2. Combine your plots
model_4_figure <- (plot_a + plot_b) + 
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title = wrapped_title, 
    theme = theme(
      plot.title = element_text(
        face = "bold", 
        size = 13,          # Slightly smaller font scales better across dual plots
        hjust = 0.5,        # Centers the text
        lineheight = 1.2,   # Adds breathing room between the title lines
        margin = margin(t = 15, b = 15) # Stronger top/bottom padding
      )
    )
  ) & 
  theme(
    legend.position = "bottom",                          
    legend.title = element_text(face = "bold", size = 10)
  ) +
  labs(color = "Race / Ethnicity")
# Display the plot
model_4_figure


#exporting t workd 
ggsave(
  filename = "Output/grace_figure_1_m4.eps", 
  plot = model_4_figure, 
  device = "eps", 
  dpi = 300,            # 300 DPI is standard print/publication quality
  width = 7.5, 
  height = 4.5, 
  units = "in",
  bg = "white" # Ensures background doesn't default to transparent
)

```

#model 5 figure 

```{r} 
#model 5 - health behaviors adjusted 

pred_grim <- predict_response(
  grim_list$M5, 
  terms = c("cancerrisk_STD [-2,-1,0,1,2]", "race_eth")
)

pred_pace <- predict_response(
  pace_list$M5, 
  terms = c("cancerrisk_STD [-2,-1,0,1,2]", "race_eth")
)

# Convert to data frames for custom ggplot handling
df_plot_grim <- as.data.frame(pred_grim)
df_plot_pace <- as.data.frame(pred_pace)

# Define matching professional palette from your example figure
color_palette <- c(
  "non-Hispanic Black" = "#377eb8",  # Clean Blue
  "non-Hispanic Other" = "#ff7f00",  # Sharp Orange
  "Hispanic"           = "#4daf4a",  # Muted Green
  "non-Hispanic White" = "#e41a1c"   # Red
)

# =========================================================================
# 2. PLOT A: GrimAge2 (Model 4 predictions)
# =========================================================================
plot_c <- ggplot(df_plot_grim, aes(x = x, y = predicted, color = group, group = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_color_manual(values = color_palette) +
  scale_x_continuous(breaks = -2:2, labels = c("-2", "-1", "0", "1", "2")) + #breaks adds ticks 
  scale_y_continuous(limits = c(-0.8, 0.8), breaks = seq(-0.5, 0.5, 0.5), labels = c("-0.5", "0.0", "+0.5")) +
  labs(
    title = "(C)",
    x = "Perceived Cancer Risk (z-score)",
    y = "Predicted GrimAge2 (z-score)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "none" # Hidden on left plot; managed globally by patchwork
  )

#dunedin pace plot for model 4 
plot_d <- ggplot(df_plot_pace, aes(
  x = x, 
  y = predicted, 
  color = group, 
  group = group
)) + geom_line(linewidth = 1.2) + 
  geom_point(size = 3) + 
  scale_color_manual(values = color_palette) + 
  scale_x_continuous(breaks = -2:2, labels = c("-2", "-1", "0", "1", "2")) + #breaks adds ticks 
  scale_y_continuous(limits = c(-0.8, 0.8), breaks = seq(-0.5, 0.5, 0.5), labels = c("-0.5", "0.0", "+0.5")) +
  labs(
    title = "(D)",
    x = "Perceived Cancer Risk (z-score)",
    y = "Predicted DunedinPACE (z-score)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "right" 
    # Hidden on left plot; managed globally by patchwork
  )


# Display the plot
model_5_figure


wrapped_title <- str_wrap(
  "Association Between Perceived Cancer Risk and Biological Aging Metrics with Regression Model 5", 
  width = 60 # Restricts lines to ~60 characters max
)

# 2. Combine your plots
model_5_figure <- (plot_c + plot_d) + 
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title = wrapped_title, 
    theme = theme(
      plot.title = element_text(
        face = "bold", 
        size = 14,          # Slightly smaller font scales better across dual plots
        hjust = 0.5,        # Centers the text
        lineheight = 1.2,   # Adds breathing room between the title lines
        margin = margin(t = 15, b = 15) # Stronger top/bottom padding
      )
    )
  ) & 
  theme(
    legend.position = "bottom",                          
    legend.title = element_text(face = "bold", size = 10) 
  ) +
  labs(color = "Race / Ethnicity")
# Display the plot
model_5_figure


#exporting t workd 
ggsave(
  filename = "Output/grace-figure_2_m5.png", 
  plot = model_5_figure, 
  device = "png", 
  dpi = 300,            # 300 DPI is standard print/publication quality
  width = 7.5, 
  height = 4.5, 
  units = "in",
  bg = "white"          # Ensures background doesn't default to transparent
)
```

