---
title: "R Notebook"
output: html_notebook
authors: Ria Baipai, Rafaela Drake, Grace McNally
date: "2026-06-11"
---


This analysis file starts from the cleaned analytic dataset created for you:

```{text}
Data/midus_cancer_aging_analysis_dataset.rds
```

Week 1 Tasks:

1. Install R and RStudio
2. Install the following R packages: tidyverse, gtsummary, flextable, broom, officer, mice, corrr, tibble, misty
3. Create ‘Table 1’ univariate statistics on the study population (put this in a Word document)


# Set Up

```{r}
# Load needed packages
library(gtsummary)
library(flextable)
library(broom)
library(officer)
library(mice)
library(corrr)
library(tibble)
library(misty)
library(haven)
library(dplyr)
library(readr)
library(tidyverse)
library(labelled)
library(Hmisc)
```


```{r}
# theme for tables
theme_gtsummary_journal(journal = "jama")
```


The main exposure is standardized perceived cancer risk:

```text
cancerrisk_STD
```

The two outcomes are:

```text
grimage2_STD
dunedinpace_STD
```

To get you started;

# Load analytic dataset

```{r load-data}
# Loading the data
midus_data <- readRDS("Data/midus_cancer_aging_analysis_dataset.rds")

# viewing data structure
glimpse(midus_data)

```



# Create ‘Table 1' univariate statistics on the study population

```{r}
# creating overall descriptive statistics
table1 <- midus_data %>% 
  select(age, sex, race_eth, marital, education, income, num_chron, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%

  tbl_summary( 
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(race_eth, sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      race_eth ~ "Race/Ethnicity",
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      num_chron ~ "Number of chronic Conditions",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table I. Univariate Statistics on Study Population")

table1

table1 %>% 
  as_flex_table() %>%
  save_as_docx(path = "Output/Table 1 Overall Descriptives.docx")

```




# Create ‘Table 2' bivariate statistics on the study population
```{r}
# Generate a basic summary table

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, num_chron, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%

tbl_summary(by = race_eth,
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      num_chron ~ "Number of chronic Conditions",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table 2: Descriptive Statistics by Race/Ethnicity on Study Population")

table2

table2 %>% 
  as_flex_table() %>%
  save_as_docx(
    path = "Output/Table 2 Descriptives by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```

# Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)

```{r}

library(tidyverse)
library(gtsummary)
library(gt)


table3 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, num_chron, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%

tbl_summary(by = race_eth,
   type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, num_chron, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical"),

    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),

    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 

    #hiding missing values??
    #applying clear labels 
    
    #do we want to use all these variables?? 
   label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      num_chron ~ "Number of chronic Conditions",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>% 
  add_overall() %>% 
  add_p() %>%
  add_stat_label() %>%
  bold_labels() %>%
  modify_caption("Table 3: Descriptive Statistics Overall and Stratified by Race and Ethnicity")

table3

table3 %>% 
  as_flex_table() %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  save_as_docx(
    path = "Output/Table 3 Descriptive Overall and by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )
```


```{r}
library(corrr)

#Table 3: Zero-Order Correlation Matrix
#Create Table 3: zero-order correlations among age, perceived cancer risk z-score, GrimAge2, DunedinPACE 
#In meeting more variables were discussed* confirm for matrix

#Correlation Matrix -> perceived cancer risk unadjusted relationships between variables with emphasis on outcomes and exposure ()

#using corrr package to create correlation matrix
Correlation_Matrix <- vars %>%
  correlate(method = "pearson", use = "pairwise.complete.obs") %>%
  shave()

# Rename all columns except term to numbers to avoid repition of variable names
names(Correlation_Matrix)[-1] <- as.character(1:6)
print(Correlation_Matrix, digits = 3)

#formatting changes (showing 3 digits for p-values)
Correlation_Matrix_fmt <- Correlation_Matrix %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "", formatC(., format = "f", digits = 3))))

flextable(Correlation_Matrix_fmt) %>%
  autofit() %>%
  save_as_docx(path = "Table_4_Correlation_Matrix.docx")
```

```{r}

# 1. LIBRARY LOADING & ENVIRONMENT SETUP
# ==============================================================================

#install.packages("VIM")

# Load required packages
library(tidyverse)    # Collection of packages for data wrangling and visualization
library(mice)         # Functions for Multivariate Imputation by Chained Equations
library(missRanger)   # Fast random forest imputation utility
library(VIM)          # Visualization and Imputation of Missing Values
library(dplyr)        # Explicitly loading dplyr (included in tidyverse)

# Set seed to ensure the random number generation (and results) are reproducible
set.seed(0864)

# Define working directory paths for file management
project_dir   <- getwd()
data_dir      <- file.path(project_dir, "Data")
output_dir    <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")

# ==============================================================================
# 2. DATA SUBSETTING & VARIABLE TYPE COERCION
# ==============================================================================

# Subset the dataset to include specific continuous outcomes/exposures and categorical covariates
midus_vars <- midus_data[ , c("grimage2_STD", "dunedinpace_STD", "cancerrisk_STD",
                             "race_eth", "age", "sex", "education", "income",
                             "smoking", "bmi", "marital", "famcancer")] %>% 
  # Explicitly convert categorical variables into factors for proper mice package handling
  mutate(race_eth  = factor(race_eth), 
        age        = factor(age),
        sex        = factor(sex),
        education  = factor(education, ordered = TRUE), # Set education as an ordered factor
        income     = factor(income), 
        smoking    = factor(smoking), 
        bmi        = factor(bmi), 
        marital    = factor(marital), 
        famcancer  = factor(famcancer))

# Define a character vector of the specific baseline confounding variables to evaluate/impute
confounders <- confounders <- c("age", "sex", "education", "income",
  "smoking", "bmi", "famcancer", "marital")

# ==============================================================================
# 3. MISSING DATA DIAGNOSTICS & EXPLORATION
# ==============================================================================

# Display the missing data pattern matrix (shows combinations of missing/observed fields)
md.pattern(midus_data) # Counting number of NA

# Display pairs of missing/observed variables to find co-missingness patterns
md.pairs(midus_data) 

# Custom function to calculate the percentage of missing values per specified variable
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars),            # Apply function across all specified target variables
        ~ mean(is.na(.x)) * 100  # Calculate percentage of NA values
      )
    ) %>% 
    # Reshape data from wide to long format for easier reading/plotting
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}

# Run the missingness summary function on the baseline midus_data for the confounders
missing_summary(midus_data, confounders)

# ==============================================================================
# 4. IMPUTATION CONFIGURATION (PREDICTOR MATRIX & METHODS)
# ==============================================================================

# Dry-run mice with 0 iterations to generate the default predictor tracking matrix
pred <- mice(midus_vars, maxit = 0)$predictorMatrix 

# Create a default imputation methods vector based on the dataframe structure
midus_methods <- make.method(midus_vars[confounders]) 
midus_methods

# Manually override and assign specific imputation algorithms for categorical variables
midus_methods["education"] <- "polyreg" # Polytomous logistic regression (ordered/unordered factors)
midus_methods["famcancer"] <- "logreg"  # Logistic regression (binary variables)
midus_methods["marital"]   <- "polyreg" # Polytomous logistic regression (unordered factors)

# ==============================================================================
# 5. DATA PARTITIONING & MULTIPLE IMPUTATION RUN
# ==============================================================================

# Isolate only the confounding variables slated for imputation
to_imp <- midus_data %>% 
  select(all_of(confounders))
to_imp 

# Isolate all remaining variables that should be preserved without change
keep <- midus_data %>%
  select(-all_of(confounders)) 
keep  

# Execute the Multivariate Imputation by Chained Equations (MICE)
imp <- mice(
  to_imp,                # The targeted dataframe containing variables to impute
  m = 20,                # Number of unique imputed datasets to generate
  seed = 0864,           # Local seed for imputation routine consistency
  printFlag = FALSE,     # Suppress the console printout history of iterations
  method = midus_methods # Use our custom-defined variable imputation methods
)

# ==============================================================================
# 6. POST-IMPUTATION RECONSTRUCTION & VERIFICATION
# ==============================================================================

# Extract the very first completely filled dataset out of the 20 generated
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Column-bind the untouched 'keep' variables back together with the newly imputed variables
df_mice_imputed <- keep %>%
  bind_cols(df_mice_completed)

# Re-run missing summary on the new dataset to verify the missingness dropped to 0%
missing_summary(df_mice_imputed, confounders)

# ==============================================================================
# 7. OPTIONAL: DIAGNOSTIC COMPARISON (PRE vs. POST IMPUTATION)
# ==============================================================================

# # Combine pre-imputation and post-imputation missing metrics for side-by-side comparison
# missing_compare <- bind_rows(
#   missing_summary(midus_data, confounders) %>%
#     mutate(dataset = "Original midus data"),
#   
#   missing_summary(df_mice_imputed, confounders) %>%
#     mutate(dataset = "Midus after mice imputation"))

# # View final structural summary of data missingness before and after processing
# missing_compare %>%
#   select(dataset, variable, percent_missing)
```
