---
title: "R Notebook"
output: html_notebook
authors: Ria Baipai, Rafaela Drake, Grace McNally
date: "2026-06-11"
---


This analysis file starts from the cleaned analytic dataset created for you:

```{text}
Data/midus_cancer_aging_analysis_dataset.rds
```

Week 1 Tasks:

1. Install R and RStudio
2. Install the following R packages: tidyverse, gtsummary, flextable, broom, officer, mice, corrr, tibble, misty
3. Create ‘Table 1’ univariate statistics on the study population (put this in a Word document)


# Set Up

```{r}
# Load needed packages
library(gtsummary)
library(flextable)
library(broom)
library(officer)
library(mice)
library(corrr)
library(tibble)
library(misty)
library(haven)
library(dplyr)
library(readr)
library(tidyverse)
library(labelled)
library(Hmisc)
library(modelsummary)
library(naniar) #mcar test
# theme for tables
theme_gtsummary_journal(journal = "jama")
```


```{r}
# theme for tables
theme_gtsummary_journal(journal = "jama")
```


The main exposure is standardized perceived cancer risk:

```text
cancerrisk_STD
```

The two outcomes are:

```text
grimage2_STD
dunedinpace_STD
```

To get you started;

# Load analytic dataset

```{r load-data}
# Loading the data
midus_data <- readRDS("Data/midus_cancer_aging_analysis_dataset.rds")

# viewing data structure
glimpse(midus_data)

```








# Create ‘Table 1' univariate statistics on the study population

```{r}
# creating overall descriptive statistics
table1 <- midus_data %>% 
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
  tbl_summary( 
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(race_eth, sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p})" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss})",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = list(all_continuous() ~ c(1,1), 
                  all_categorical() ~ c(0,1)
    ),
    label = list(
      race_eth ~ "Race/Ethnicity",
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table I. Univariate Statistics on Study Population")

table1

table1 %>% 
  as_flex_table() %>%
  save_as_docx(path = "Output/Grace_Table 1 Overall Descriptives.docx")

```
#Table 2 

# Create ‘Table 2' bivariate statistics on the study population
```{r}
# Generate a basic summary table

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
tbl_summary(by = race_eth,
     type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical") ,
    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p}%)" 
    ),
    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss})",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 
    label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>%
  bold_labels() %>%
  modify_caption("Table 2: Descriptive Statistics by Race/Ethnicity on Study Population")

table2

table2 %>% 
  as_flex_table() %>%
  save_as_docx(
    path = "Output/Grace_Table 2 Descriptives by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```


# Table 3: 

```{r}

library(tidyverse)
library(gtsummary)
library(gt)



table3 <- midus_data %>%
  select(age, sex, race_eth, marital, education, income, famcancer, efficacy, 
         smoking, alcohol, bmi, cancerrisk, grimage2, dunedinpace) %>%
 filter(
    !is.na(grimage2) #dropping missing value from grimage2
  ) %>%
tbl_summary(by = race_eth,
   type = list(
      # Quantitative (Continuous) Variables
      c(grimage2, dunedinpace, bmi, age, income, cancerrisk, efficacy) ~ "continuous",
      # Categorical Variables
      c(sex, marital, education, smoking, alcohol, famcancer) ~ "categorical"),

    statistic = list( 
      all_continuous() ~ "{mean} ({sd})",   # Removed the c() wrapper
      all_categorical() ~ "{n} ({p})" 
    ),

    missing = "ifany",
    missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
    missing_text = "Missing", 
    digits = all_continuous() ~ 1, 

    #hiding missing values??
    #applying clear labels 
    
    #do we want to use all these variables?? 
   label = list(
      grimage2 ~ "GrimAge2", 
      dunedinpace ~ "DunedinPACE", 
      bmi ~ "Body mass index (kg/m²)",
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital status",
      education ~ "Education attainment",
      income ~ "Annual household income",
      smoking ~ "Smoking status",
      alcohol ~ "Alcohol consumption",
      famcancer ~ "Family cancer history",
      cancerrisk ~ "Perceived cancer risk", 
      efficacy ~ "Self-efficacy for cancer prevention"))  %>% 
  add_overall() %>% 
      add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list( 
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  add_stat_label() %>%
  bold_labels() %>%
  modify_caption("Table 3: Descriptive Statistics Overall and Stratified by Race and Ethnicity")

table3

table3 %>% 
  as_flex_table() %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  save_as_docx(
    path = "Output/Grace_Table 3 Descriptive Overall and by Race and Ethnicity.docx",
    pr_section = prop_section(
      page_size = page_size(orient = "landscape", width = 11.7, height = 8.3))
      )

```


#Table 4: correlation matrix

```{r}
#Table 4: Zero-Order Correlation Matrix -> perceived cancer risk unadjusted relationships between variables with #emphasis on outcomes and exposure
cor_vars <- midus_data %>%
  select(grimage2, dunedinpace, age, cancerrisk_STD, efficacy) %>%
 filter(  !is.na(grimage2) #dropping missing value from grimage2
  )
#Convert to a matrix
vars_matrix <- as.matrix(cor_vars)

#Correlation matrix using Hmisc
correlation_results <- Hmisc::rcorr(vars_matrix, type=c("pearson"))
#extract correlation matrix
r_matrix <- correlation_results$r
#extract p-values
p_matrix <- correlation_results$P

#adding starts for p-values and keep lower triagnle
r_formatted <- format(round(r_matrix, 3), nsmall = 3) #issue
stars <- ifelse(p_matrix <0.001, "***",
                ifelse(p_matrix < 0.01, "**",
                       ifelse(p_matrix <0.05, "*", "")))
#combined
final_cor_matrix <- matrix(paste0(r_formatted, stars), nrow = nrow(r_matrix))
rownames(final_cor_matrix) <- rownames(r_matrix)
colnames(final_cor_matrix) <- colnames(r_matrix)

#remove self-correlations
diag(final_cor_matrix) <- ""

#only keep lower triangle
final_cor_matrix[upper.tri(final_cor_matrix)] <- ""

#convert to df and format row names as a separate column
final_df <- as.data.frame(final_cor_matrix)
final_df <- tibble::rownames_to_column(final_df, var = "Variable")

#Professional variable labels
#vector mapping old names to new labels
clean_labels <- c(
  "grimage2"        = "GrimAge2 z-score",
  "dunedinpace"     = "DunedinPACE z-score",
  "age"             = "Age",
  "cancerrisk_STD"  = "Perceived cancer risk z-score",
  "efficacy"        = "Self-efficacy for Cancer Prevention"
)

#rename in final df
final_df$Variable <- clean_labels[final_df$Variable]
#add numbers to row names
final_df$Variable <- paste0(1:nrow(final_df), ". ", final_df$Variable) 
#change column headers to numbers (corresponding with row numbering)
colnames(final_df)[2:ncol(final_df)] <- 1:(ncol(final_df) - 1)


#formatting (flextable)
ft <- flextable(final_df) %>%
  #set header labels
  set_header_labels(
    Variable = "Variable",
    grimage2 = "1",
    dunedinpace = "2",
    age = "3",
    num_chron = "4",
    cancerrisk_STD = "5",
    efficacy = "6") %>%
  theme_booktabs() %>% #APA
  bold(part = "header") %>%
  autofit() %>%
  fit_to_width(max_width = 6.5, unit = "in") %>%
  align(j = 1, align = "left", part = "all") %>% 
  align(j = 2:ncol(final_df), align = "center", part = "all") %>%
  # Add a table title
  add_header_lines("1.1. Table 3. Zero-order correlations among epigenetic aging measures, perceived cancer risk, age, self-efficacy, and chronic conditions.") %>%
  #footer with stars info
  add_footer_lines("* p < 0.05, ** p < 0.01, *** p < 0.001") %>%
  font(fontname = "Arial", part = "all")

ft

#export
save_as_docx(ft, path = "Output/Grace_Table_4_Correlation_Matrix.docx")
```



# Multiple Imputation


```{r}
# Conducting multiple imputation 

# Subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates
#15 vars total 
# checking for row containing NA value in grimAge2_STD
print(is.na(midus_data$grimage2_STD))

midus_vars <- midus_data %>% 
              select("income", "famcancer", "efficacy",
                     "marital", "education", "grimage2_STD", "dunedinpace_STD",
                     "cancerrisk_STD", "race_eth", "age", "sex",
                     "smoking", "alcohol", "bmi") %>%
  # Removing row with NA value for grimage2_STD
  filter(
    !is.na(grimage2_STD) #dropping missing value from grimage2
  ) %>%
  # Categorical variables should be coded as factors in order to invoke polyreg
  mutate(
      race_eth = factor(race_eth),
      alcohol = factor(alcohol),
      sex = factor(sex),
      education = factor(education, ordered = TRUE),
      smoking = factor(smoking),
      marital = factor(marital),
      famcancer = factor(famcancer)
    )


# confirming NA value has been removed
print(is.na(midus_vars$grimage2_STD))

# Run Little's MCAR
library(naniar)  # you need this package to run MCAR (Roph)
mcar_test(midus_vars)
# statistic df  p-value  missing.patterns
# 212.3461	162	0.004844206	13	

# Create a clean character vector tracking all variables in our target analysis subset
analysis_variables <- names(midus_vars)

md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns 
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

#missing summary function 
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    ) %>%
  arrange(desc(percent_missing)) #arrange output in descending over 
}


missing_summary(midus_data, analysis_variables)

#predictor matrix 
pred <- mice(midus_vars, maxit = 0)$predictorMatrix 

# Creating a methods object
midus_methods <- make.method(midus_vars) 
midus_methods # Looking to check method types for categorical vars

# Assigning new methods to categorical variables
midus_methods["education"] <- "polr" # changed this from "polyreg" as education is an ordered categorical variable
#polyreg used for categorical values with 2+ categories
midus_methods["famcancer"] <- "logreg" #logreg for nominal categorical values
midus_methods["marital"] <- "polyreg"
midus_methods # Checking to ensure variables are updated

imp <- mice(
  midus_vars, # data: what we want to impute
  m = 10, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed


# Re-run missing summary on the complete dataset to verify all missingness dropped to 0%
missing_summary(df_mice_completed, analysis_variables)

```


#regression analysis

```{r}
# =========================================
#Regression analysis 
# Split into two explicit data frames as requested
df_overall <- df_mice_completed
#df_white   <- df_mice_completed %>% filter(race_eth == "non-Hispanic White")
#df_white
# --- Outcome: GrimAge2 (Overall) ---
# --- Outcome: GrimAge (Overall) ---

# Model 1: perceived cancer risk + age + sex + race/ethnicity + marital status + education + income
grim_ov_m1 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income, 
                 data = df_overall)

# Model 2: Model 1 + family cancer history
grim_ov_m2 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer, 
                 data = df_overall)

# Model 3: Model 2 + self-efficacy for cancer prevention
grim_ov_m3 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy, 
                 data = df_overall)

# Model 4: Model 3 + smoking + alcohol use
grim_ov_m4 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol, 
                 data = df_overall)

# Model 5: Model 4 + BMI
grim_ov_m5 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol + bmi, 
                 data = df_overall)

# Model 6: Model 5 + perceived cancer risk × race/ethnicity interaction
grim_ov_m6 <- lm(grimage2_STD ~  age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol + bmi + (cancerrisk_STD * race_eth), 
                 data = df_overall) #cancer time race

#outcome dunedinpace

# Model 1: perceived cancer risk + age + sex + race/ethnicity + marital status + education + income
pace_ov_m1 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income, 
                 data = df_overall)

# Model 2: Model 1 + family cancer history
pace_ov_m2 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer, 
                 data = df_overall)

# Model 3: Model 2 + self-efficacy for cancer prevention
pace_ov_m3 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy, 
                 data = df_overall)

# Model 4: Model 3 + smoking + alcohol use
pace_ov_m4 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol, 
                 data = df_overall)

# Model 5: Model 4 + BMI
pace_ov_m5 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol + bmi, 
                 data = df_overall)

# Model 6: Model 5 + perceived cancer risk × race/ethnicity interaction
pace_ov_m6 <- lm(dunedinpace_STD ~ + age + sex + race_eth + marital + education + income + 
                 famcancer + efficacy + smoking + alcohol + bmi + (cancerrisk_STD * race_eth), 
                 data = df_overall)

# --- Outcome: GrimAge2 (White Sample) ---
#grim_wh_m1 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital, data = df_white)
#grim_wh_m2 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital + education + income, data = df_white)
#grim_wh_m3 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer, data = df_white)
#grim_wh_m4 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer + efficacy, data = df_white)
#grim_wh_m5 <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer + efficacy + smoking + alcohol + bmi, data = df_white)

# --- Outcome: DunedinPACE (White Sample) ---
#pace_wh_m1 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + marital, data = df_white)
#pace_wh_m2 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + marital + education + income, data = df_white)
#pace_wh_m3 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer, data = df_white)
#pace_wh_m4 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer + efficacy, data = df_white)
#pace_wh_m5 <- lm(dunedinpace_STD ~ cancerrisk_STD + age + sex + marital + education + income + num_chron + famcancer + efficacy + smoking + alcohol + bmi, data = df_white)

#summary(pace_wh_m1)
summary(pace_ov_m1)

#saving to 


# moel objects into named list
grim_list <- list(
  "M1" = grim_ov_m1, "M2" = grim_ov_m2, "M3" = grim_ov_m3,
  "M4" = grim_ov_m4, "M5" = grim_ov_m5, "M6" = grim_ov_m6
)

pace_list <- list(
  "M1" = pace_ov_m1, "M2" = pace_ov_m2, "M3" = pace_ov_m3,
  "M4" = pace_ov_m4, "M5" = pace_ov_m5, "M6" = pace_ov_m6
)

#what we want to show



  # modelsummary to automatically extract Estimates, CIs, and p-values ---
df_grim <- modelsummary(
  grim_list, 
  output    = "data.frame", 
  estimate  = "{estimate} {stars}",
  statistic = "[{conf.low}, {conf.high}]", # <-- Force 95% CI beneath the estimate
  stars     = TRUE,  # Keeps the significance stars (*, **, ***) on the estimate
  fmt       = 3,     # Three decimal places
 coef_map = c("cancerrisk_STD" = "GrimAge2"),
  gof_omit = ".*",
  #gof_map   = c("nobs", "r.squared") # number of obs and R^2 at the bottom of the table
) 

df_pace <- modelsummary(
  pace_list, 
  output    = "data.frame", 
  estimate  = "{estimate} {stars}",
  statistic = "[{conf.low}, {conf.high}]", # <-- Force 95% CI beneath the estimate
  stars     = TRUE,  
  fmt       = 3, 
  coef_map  = c("cancerrisk_STD" = "DunedinPACE"), 
  gof_omit = ".*"
  #gof_map   = c("nobs", "r.squared")
) 

# FIXED: Changed the second df_grim to df_pace so both models stack
stacked_df <- bind_rows(df_grim, df_pace) %>% 
  select(-part, -statistic) 

# 4. Turn the stacked data frame into your final flextable and format headers
final_table_flextable <- flextable(stacked_df) %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  # Remove the duplicate labels in the first column ("term") by blanking out rows 2 and 4
  mk_par(i = c(2, 4), j = "term", value = as_paragraph("")) %>%
  # Remove the "term" header text by replacing it with a blank space ""
  set_header_labels(
    term = "",
    M1 = "M1", M2 = "M2", M3 = "M3", M4 = "M4", M5 = "M5", M6 = "M6"
  ) %>%
  
  # Add the β (95% CI) subheader row directly beneath M1 - M6
  add_header_row(
    values = c("", rep("β (95% CI)", 6)),
    top = FALSE # Tells flextable to place this row below the M1-M6 row
  ) %>%
  
  
  # Make the M1 - M6 headers bold (row 1 of the header)
  bold(i = 1, j = 2:7, part = "header") %>%
  
  # Make the row labels bold (only rows 1 and 3 since 2 and 4 are now blank)
  bold(i = c(1, 3), j = 1, part = "body") %>%
  
  # Clean up alignments for professional formatting
  align(align = "center", part = "header") %>%
  align(j = 1, align = "left", part = "header")

# 5. Save the final flextable to Word
# FIXED: Re-aligned the backticks and quotes for the caption name block
save_as_docx(
  `Table 4: Overall models with standardized beta coefficients (and 95% confidence intervals) from a series of linear regression models quantifying the association between perceived cancer risk and epigenetic aging among MIDUS II sample ` = final_table_flextable,  
  path = "Output/grace_overall_models.docx",
  pr_section = prop_section(
    page_size = page_size(orient = "landscape", width = 11.7, height = 8.3)
  )
)


```

