---
title: "PQAR 2026 Research Project"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing necessary R packages
```{r}
<<<<<<< HEAD
  library("gtsummary")
  library("flextable")
  library("broom")
  library("officer")
  library("mice")
  library("corrr")
  library("tibble")
  library("misty")
  library("haven")
  library("dplyr")
  library("Hmisc")
```
Classification of variables:
- Demographic information: age, sex, race/ethnicity, marital status, education, income
- Physical considerations: smoking, number of chronic diseases, BMI, alcohol, family cancer history
- Psychological considerations: efficacy
- Biological data: dunedinpace, grimage2
- Dependent variable of this study: cancer risk


Reading in data and creating new dataframe based on demographic characteristics
```{r}
  project_dir <- getwd()
  data_dir  <- file.path(project_dir, "Data")
  output_dir  <- file.path(project_dir, "Output")
  analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
  library(dplyr)
  Canc_data <- readRDS(analysis_path) #Canc_data: data from MIDUS cancer subproject
  
  # Setting variables for data table
  age <- Canc_data$age 
  sex <- Canc_data$sex 
  edu <- Canc_data$education 
  income <- Canc_data$income 
  race_ethn <- Canc_data$race_eth 
  mar_stat <- Canc_data$marital
  bmi <- Canc_data$bmi
  grimage2 <- Canc_data$grimage2_STD
  dunedinpace <- Canc_data$dunedinpace_STD
  num_chron <- Canc_data$num_chron
  smoking <- Canc_data$smoking
  alcohol <- Canc_data$alcohol
  famcancer <- Canc_data$famcancer
  cancerrisk <- Canc_data$cancerrisk
  self_efficacy <- Canc_data$efficacy
  
  
  # Create dataframe based on given variables
  df1 <- data.frame(age, race_ethn, sex, mar_stat, edu, income, bmi, grimage2, dunedinpace, num_chron, smoking, alcohol, famcancer, cancerrisk, self_efficacy)
  
# Adding descriptive labels to data frame
  library(labelled)
  var_label(df1$age) <- "Age"
  var_label(df1$sex) <- "Sex"
  var_label(df1$mar_stat) <- "Marital Status"
  var_label(df1$edu) <- "Education"
  var_label(df1$income) <- "Income"
  var_label(df1$bmi) <- "BMI"
  var_label(df1$race_ethn) <- "Race and Ethnicity"
  var_label(df1$grimage2) <- "GrimAge2"
  var_label(df1$dunedinpace) <- "DunedinPACE"
  var_label(df1$num_chron) <- "Number of Chronic Conditions"
  var_label(df1$smoking) <- "Smoking Status"
  var_label(df1$alcohol) <- "Alcohol Consumption"
  var_label(df1$famcancer) <- "Family Cancer History"
  var_label(df1$self_efficacy) <- "Self Efficacy"
  var_label(df1$cancerrisk) <- "Perceived Cancer Risk"
  colnames(df1) <- Hmisc::label(df1) # Assigning new labels to data frame
  glimpse(df1) 
=======
  install.packages("gtsummary")
  install.packages("flextable")
  install.packages("broom")
  install.packages("officer")
  install.packages("mice")
  install.packages("corrr")
  install.packages("tibble")
  install.packages("misty")
  install.packages("haven")
  install.packages("dplyr")
```


# Reading in data and creating new dataframe based on demographic characteristics
```{r}
  library(dplyr)
  Canc_data <- readRDS("~/Desktop/Research_Project_PQAR/midus_cancer_aging_analysis_dataset.rds") #Canc_data: data from MIDUS cancer subproject
  age <- Canc_data$age # Should we use both age and birth year? Just use Age --> Rophence
  race_ethn <- Canc_data$race_eth # Participant race and ethnicity
  sex <- Canc_data$sex # Participant sex
  mar_stat <- Canc_data$marital # Marital status
  edu <- Canc_data$education # Education level
  income <- Canc_data$income # Income
  
 # pt_id <- Canc_data$id # Participant MIDUS ID
  
  
  # Create dataframe based on given variables (Make sure you use all 14 variables as listed in the Dataset section of your syllabus -->Rophence)
  df1 <- data.frame(Age = age, Race_Ethnicity = race_ethn, Sex = sex, Marital_Status = mar_stat, Education = edu, Income = income)
  df1
>>>>>>> dc3ed2ed5e508b2526bd70b1e1fcd7a54493bf35
```


Create ‘Table 1' univariate statistics on the study population
```{r}
# Generate a basic summary table
library(gtsummary)
library(tidyverse)
theme_gtsummary_journal(journal = "jama")
<<<<<<< HEAD
table1 <-
  df1 %>% 
  select(-`Race and Ethnicity`) %>% # REMOVING race&ethn to make merging (table 3) easier
  tbl_summary( 
    statistic = 
      list( 
        all_continuous() ~ c("{mean} ({sd})"),
        all_categorical() ~ "{n} ({p}%)" 
        ), 
     missing = "no",
            ) %>%
  modify_caption("**Table I. Univariate Demographic Characteristics of MIDUS Participants**") 

  table1

```


Table 2
=======
df1 %>% # Relabel unknown --> missing, add mean instead of median, use label function to eliminate underscores/clean titles 
  tbl_summary(missing_text = "Missing") %>% 
  modify_caption("**Table I. Univariate Demographic Characteristics of MIDUS Participants**")

```

# Table 1 info cont.
```{r}
# Create a histogram
histogram_plot <- df1 %>%
  ggplot(aes(x = income)) +
  geom_histogram(bins = 30, color = "white") +
  labs(title = "Distribution of Income",
       x = "Income", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # this centers the title

# Show the plot in the notebook
histogram_plot

# Save the histogram as a PNG image
ggsave("example_income_histogram.png", 
       plot = histogram_plot, width = 7, height = 5, dpi = 300)
```


# Table 2
>>>>>>> dc3ed2ed5e508b2526bd70b1e1fcd7a54493bf35
```{r}
library(gtsummary)
library(tidyverse)
# Generate a basic summary table
theme_gtsummary_journal(journal = "jama")
<<<<<<< HEAD
table2 <- 
  df1 %>% # add p-value column - cmd is add p / simulation of 2000 replicates
  tbl_summary(
  by = `Race and Ethnicity`,
  statistic = 
        list( 
          all_continuous() ~ c("{mean} ({sd})"),
          all_categorical() ~ "{n} ({p}%)" 
          ), 
       missing_text = "Missing"
               ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table II. Demographic Characteristics of MIDUS Participants by Race and Ethnicity**")

table2
  
```

## Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)
```{r}
  age <- Canc_data$age 
  sex <- Canc_data$sex 
  edu <- Canc_data$education 
  income <- Canc_data$income 
  race_ethn <- Canc_data$race_eth 
  mar_stat <- Canc_data$marital
  bmi <- Canc_data$bmi
  grimage2 <- Canc_data$grimage2_STD
  dunedinpace <- Canc_data$dunedinpace_STD
  num_chron <- Canc_data$num_chron
  smoking <- Canc_data$smoking
  alcohol <- Canc_data$alcohol
  famcancer <- Canc_data$famcancer
  cancerrisk <- Canc_data$cancerrisk
  self_efficacy <- Canc_data$efficacy
  
library(gtsummary)
# Merging both tables above: issue with a lot of white space underneath
combined_table <- tbl_merge(
  tbls = list(table1, table2),
  tab_spanner = c("**Overall**", NA)
)

combined_table
```

# Export as word document
=======
df1 %>% # add p-value column - cmd is add p / simulation of 2000 replicates
  tbl_summary(
  by = Race_Ethnicity
  ) %>%
  modify_caption("**Table II. Demographic Characteristics of MIDUS Participants by Race and Ethnicity**") # or table I ? 
```

## Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)
>>>>>>> dc3ed2ed5e508b2526bd70b1e1fcd7a54493bf35

