---
title: "PQAR 2026 Research Project"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing necessary R packages
```{r}
  # library("gtsummary")
  # library("flextable")
  # library("broom")
  # library("officer")
  # library("mice")
  # library("corrr")
  # library("tibble")
  # library("misty")
  # library("haven")
  # library("dplyr")
  # library("Hmisc")
```
Classification of variables:
- Demographic information: age, sex, race/ethnicity, marital status, education, income
- Physical considerations: smoking, number of chronic diseases, BMI, alcohol, family cancer history
- Psychological considerations: efficacy
- Biological data: dunedinpace, grimage2
- Dependent variable of this study: cancer risk


# Reading in data and creating new dataframe based on demographic characteristics
```{r}
  project_dir <- getwd()
  data_dir  <- file.path(project_dir, "Data")
  output_dir  <- file.path(project_dir, "Output")
  analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
  library(dplyr)
  Canc_data <- readRDS(analysis_path) #Canc_data: data from MIDUS cancer subproject
  
  # Setting variables for data table
  age <- Canc_data$age 
  sex <- Canc_data$sex 
  edu <- Canc_data$education 
  income <- Canc_data$income 
  race_ethn <- Canc_data$race_eth 
  mar_stat <- Canc_data$marital
  bmi <- Canc_data$bmi
  grimage2 <- Canc_data$grimage2_STD
  dunedinpace <- Canc_data$dunedinpace_STD
  num_chron <- Canc_data$num_chron
  smoking <- Canc_data$smoking
  alcohol <- Canc_data$alcohol
  famcancer <- Canc_data$famcancer
  cancerrisk <- Canc_data$cancerrisk
  self_efficacy <- Canc_data$efficacy
  
  
  # Create dataframe based on given variables
  df1 <- data.frame(age, race_ethn, sex, mar_stat, edu, income, bmi, grimage2, dunedinpace, num_chron, smoking, alcohol, famcancer, cancerrisk, self_efficacy)
  
# Adding descriptive labels to data frame
  library(labelled)
  var_label(df1$age) <- "Age"
  var_label(df1$sex) <- "Sex"
  var_label(df1$mar_stat) <- "Marital Status"
  var_label(df1$edu) <- "Education"
  var_label(df1$income) <- "Income"
  var_label(df1$bmi) <- "BMI"
  var_label(df1$race_ethn) <- "Race and Ethnicity"
  var_label(df1$grimage2) <- "GrimAge2"
  var_label(df1$dunedinpace) <- "DunedinPACE"
  var_label(df1$num_chron) <- "Number of Chronic Conditions"
  var_label(df1$smoking) <- "Smoking Status"
  var_label(df1$alcohol) <- "Alcohol Consumption"
  var_label(df1$famcancer) <- "Family Cancer History"
  var_label(df1$self_efficacy) <- "Self Efficacy"
  var_label(df1$cancerrisk) <- "Perceived Cancer Risk"
  colnames(df1) <- Hmisc::label(df1) # Assigning new labels to data frame
  glimpse(df1) 

```



Create ‘Table 1' univariate statistics on the study population
```{r}
# Generate a basic summary table
library(gtsummary)
library(tidyverse)
theme_gtsummary_journal(journal = "jama")

table1 <-
  df1 %>% 
  select(-`Race and Ethnicity`) %>% # REMOVING race&ethn to make merging (table 3) easier
  tbl_summary( 
    type = list(c(Sex, `Marital Status`, Education, `Smoking Status`, 
                `Alcohol Consumption`, `Family Cancer History`) ~ "categorical",
                c(Age, Income, BMI, GrimAge2, 
                DunedinPACE, `Number of Chronic Conditions`,
                `Perceived Cancer Risk`, `Self Efficacy`) ~ "continuous"),
    statistic = 
      list( 
        all_continuous() ~ c("{mean} ({sd})"),
        all_categorical() ~ "{n} ({p}%)" 
        ), 
     missing = "no",
            ) %>%
  modify_caption("**Table I. Univariate Demographic Characteristics of MIDUS Participants**") 

  table1

```


# Table 2
```{r}
library(gtsummary)
library(tidyverse)
# Generate a basic summary table
table2_cont <- df1 %>% 
  tbl_summary(
    by = `Race and Ethnicity`,
    # Defining variables as continuous
    include = c(Age, Income, BMI, GrimAge2, 
                DunedinPACE, `Number of Chronic Conditions`,
                `Perceived Cancer Risk`, `Self Efficacy`),
    type = list(everything() ~ "continuous"), 
    # Specifying type of statistical analysis on continuous values
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    # Do not show missing values
    missing = "no"
  ) %>%
  # Calculate p-values
  add_overall() %>% 
  add_p(all_continuous() ~ "kruskal.test")


table2_cat <- df1 %>% 
  tbl_summary(
    by = `Race and Ethnicity`,
    # Defining variables as categorical
    include = c(Sex, `Marital Status`, Education, `Smoking Status`, 
                `Alcohol Consumption`, `Family Cancer History`),
    type = list(everything() ~ "categorical"), 
    # Specifying type of statistical analysis on categorical values
    statistic = list(
      all_categorical() ~ "{n} ({p}%)"
    ),
    # Indicate missing values, their count, and percentage
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  # Calculate p-values
  add_overall() %>% 
  add_p(all_categorical() ~ "chisq.test.no.correct")


# Merging continuous & categorical tables
theme_gtsummary_journal(journal = "jama")
table2 <- tbl_stack(
  tbls = list(table2_cat, table2_cont),
  )

table2
  
```

## Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)
```{r}
  age <- Canc_data$age 
  sex <- Canc_data$sex 
  edu <- Canc_data$education 
  income <- Canc_data$income 
  race_ethn <- Canc_data$race_eth 
  mar_stat <- Canc_data$marital
  bmi <- Canc_data$bmi
  grimage2 <- Canc_data$grimage2_STD
  dunedinpace <- Canc_data$dunedinpace_STD
  num_chron <- Canc_data$num_chron
  smoking <- Canc_data$smoking
  alcohol <- Canc_data$alcohol
  famcancer <- Canc_data$famcancer
  cancerrisk <- Canc_data$cancerrisk
  self_efficacy <- Canc_data$efficacy
  
library(gtsummary)
# Merging both tables above
theme_gtsummary_journal(journal = "jama")
combined_table <- tbl_merge(
  tbls = list(table2, table1),
  tab_spanner = FALSE
)

combined_table
```

# Export as word document
```{r}
  library(flextable)
  combined_table %>%
  as_flex_table() %>%
  flextable::save_as_docx(path = "table_3.docx")
```




