---
title: "PQAR 2026 Research Project"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing necessary R packages
```{r}
# COMMENT ON THESE LIBRARIES
library("gtsummary")
library("flextable")
library("broom")
library("officer")
library("mice")
library("corrr")
library("tibble")
library("misty")
library("haven")
library("dplyr")
library("Hmisc")
```


Classification of variables:
- Demographic information: age, sex, race/ethnicity, marital status, education, income
- Physical considerations: smoking, number of chronic diseases, BMI, alcohol, family cancer history
- Psychological considerations: efficacy
- Biological data: dunedinpace, grimage2
- Dependent variable of this study: cancer risk


# Reading in data and creating new dataframe based on demographic characteristics
```{r}
# Please comment your code so that someone else can follow along, for example, the code below is getting project working directory, also don't indent your code, add a line space to separate different lines of codes

# Get current project directory (RO)
project_dir <- getwd()

# Get data directory
data_dir  <- file.path(project_dir, "Data")
output_dir  <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
  
# loading in the datatset  
midus_data <- readRDS(analysis_path) #midus_data: data from MIDUS cancer subproject
  
# Adding descriptive labels to data frame >> you did not need to do this, as you are subsetting the dataset just to create labels, yet subseqent anlysis will require some of the standardized variables not included in this subset. One way to add labels is using the label function in tbl_summary. What you did here creates spaces in Variable names and variable names should not have spaces, so use label function within tbl_summary

```



Create ‘Table 1' univariate statistics on the study population


```{r}
# Generate a basic summary table
theme_gtsummary_journal(journal = "jama")


table1 <- midus_data %>%
  # Selecting for variables that will displayed in table
   select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%
  tbl_summary( 
    # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race and Ethnicity",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    # Assigning variable type (categorical or continuous)
    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer", "race_eth")) ~ "categorical",
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),
    statistic = 
      # Specifying displayed data depending on variable type
      list( 
        all_continuous() ~ "{mean} ({sd})",
        all_categorical() ~ "{n} ({p}%)" 
        ), 
     missing = "ifany"
            )
  
  table1 <- table1 %>%
  modify_caption("**Table I. Univariate Demographic Characteristics of MIDUS Participants**") 

  table1

```

```{r}
# Load packages
library(tidyverse)
library(gtsummary)

# Table 1: Overall descriptive statistics
# Do not remove Race and Ethnicity here.
# Race and Ethnicity is needed for the overall descriptive table.
# Remove it only later when creating race-stratified or merged tables.

# For R to recognize the missing values in specific variables as missing we need to explicitly create a missing category for categorical variables that have missing values

df1 <- df1 %>%
  mutate(
    `Marital Status` = as.character(`Marital Status`),
    `Marital Status` = if_else(is.na(`Marital Status`), "Missing", `Marital Status`),
    `Marital Status` = factor(
    `Marital Status`,
      levels = c(
        "Married",
        "Divorced/Separated/Widowed",
        "Never married",
        "Missing"
      )
    ),

    Education = as.character(Education),
    Education = if_else(is.na(Education), "Missing", Education),
    Education = factor(
      Education,
      levels = c(
        "High school or less",
        "Some college/associate's degree",
        "College degree or higher",
        "Missing"
      )
    ),

    `Family Cancer History` = as.character(`Family Cancer History`),
    `Family Cancer History` = if_else(is.na(`Family Cancer History`), "Missing", `Family Cancer History`),
    `Family Cancer History` = factor(`Family Cancer History`, levels = c("No", "Yes", "Missing"))
  )

table1 <- df1 %>%
  tbl_summary(include = c(`Race and Ethnicity`, Sex, `Marital Status`, Education,
                          `Smoking Status`, `Alcohol Consumption`, `Family Cancer History`,
                          Age, Income, BMI, GrimAge2, DunedinPACE,`Number of Chronic Conditions`,
                          `Perceived Cancer Risk`, `Self Efficacy`),
              type = list(c(`Race and Ethnicity`, Sex, `Marital Status`, Education,
                            `Smoking Status`, `Alcohol Consumption`, `Family Cancer History`) ~ "categorical",
                          c(Age, Income, BMI, GrimAge2, DunedinPACE, 
                            `Number of Chronic Conditions`, `Perceived Cancer Risk`, `Self Efficacy`) ~ "continuous"),
              
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
              digits = list(all_continuous() ~ 2,
                            all_categorical() ~ c(0, 1)), 
              missing = "ifany",
              missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
              missing_text = "Missing") %>%  # here you can now use  missing_text = "Missing" if you have coded categorical variables with missing values as Missing, previously it was missing = "ifany" which also shows missing values in continouous variables, and we do not want that.
  modify_caption("**Table 1. Overall Descriptive Characteristics of MIDUS Participants**") %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table1
```

# Table 2

Rophence>>> this is not the right way to go about it, you don't need to do it separately for continuous then separately for categorical, and then merge them, you just need to modify your code for Table 1 above  and use the `by` syntax within tbl_summary, so your code here is wrong


```{r}
library(gtsummary)  # once you load libraries you don't need to load them again in every R chunk
library(tidyverse)
# Generate a basic summary table
table2_cont <- df1 %>% 
  tbl_summary(
    by = `Race and Ethnicity`,
    # Defining variables as continuous
    include = c(Age, Income, BMI, GrimAge2, 
                DunedinPACE, `Number of Chronic Conditions`,
                `Perceived Cancer Risk`, `Self Efficacy`),
    type = list(everything() ~ "continuous"), 
    # Specifying type of statistical analysis on continuous values
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    # Do not show missing values
    missing = "no"
  ) %>%
  # Calculate p-values
  add_overall() %>% 
  add_p(all_continuous() ~ "kruskal.test")


table2_cat <- df1 %>% 
  tbl_summary(
    by = `Race and Ethnicity`,
    # Defining variables as categorical
    include = c(Sex, `Marital Status`, Education, `Smoking Status`, 
                `Alcohol Consumption`, `Family Cancer History`),
    type = list(everything() ~ "categorical"), 
    # Specifying type of statistical analysis on categorical values
    statistic = list(
      all_categorical() ~ "{n} ({p}%)"
    ),
    # Indicate missing values, their count, and percentage
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  # Calculate p-values
  add_overall() %>% 
  add_p(all_categorical() ~ "chisq.test") # Unclear on how p-value is being calculated for categorical values
  add_overall() %>%  # this should not have been added, it's part of what makes Table 3, Table 2 was just stratified by race
  add_p(all_categorical() ~ "chisq.test.no.correct")  # this is wrong, some variables are continuous so you need two arguments on how to find p values for continuous and for categorical


# Merging continuous & categorical tables
theme_gtsummary_journal(journal = "jama")
table2 <- tbl_stack(
  tbls = list(table2_cat, table2_cont),
  )

table2
  
```


the correct code for Table 2, would just be using the correct code for table 1 and adding in `by`, see below

```{r}
# Table 2: Descriptive statistics by Race and Ethnicity
# No overall column.
# P-values compare each variable across Race/Ethnicity groups.
# Chi-square tests are used for categorical variables to avoid Fisher exact test errors.

table2 <- df1 %>%
  tbl_summary(
    by = `Race and Ethnicity`,

    include = c(Sex, `Marital Status`, Education,
                `Smoking Status`, `Alcohol Consumption`,
                `Family Cancer History`, Age, Income, BMI,
                GrimAge2, DunedinPACE,
                `Number of Chronic Conditions`,
                `Perceived Cancer Risk`, `Self Efficacy`),

    type = list(
      c(Sex, `Marital Status`, Education,
        `Smoking Status`, `Alcohol Consumption`,
        `Family Cancer History`) ~ "categorical",
      c(Age, Income, BMI, GrimAge2, DunedinPACE,
        `Number of Chronic Conditions`,
        `Perceived Cancer Risk`, `Self Efficacy`) ~ "continuous"
    ),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "**Table 2. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity**"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table2
```


Running the above gives you warnings, never ignore warnings because they always tell you something is not right. The warning means the regular chi-square approximation may not be reliable because some Race/Ethnicity × category cells are small. Instead of forcing Fisher’s exact test, which caused the FEXACT workspace error, use Monte Carlo simulation for the chi-square p-value. Thus, for categorical variables, use chi-square tests with simulated p-values. This avoids Fisher exact test errors and reduces problems caused by small cell counts.


Thus correct code for cretaing table 2 will be as follows:

```{r}
table2 <- df1 %>%
  tbl_summary(
    by = `Race and Ethnicity`,

    include = c(Sex, `Marital Status`, Education,
                `Smoking Status`, `Alcohol Consumption`,
                `Family Cancer History`, Age, Income, BMI,
                GrimAge2, DunedinPACE,
                `Number of Chronic Conditions`,
                `Perceived Cancer Risk`, `Self Efficacy`),

    type = list(
      c(Sex, `Marital Status`, Education,
        `Smoking Status`, `Alcohol Consumption`,
        `Family Cancer History`) ~ "categorical",
      c(Age, Income, BMI, GrimAge2, DunedinPACE,
        `Number of Chronic Conditions`, `Perceived Cancer Risk`,
        `Self Efficacy`) ~ "continuous"
    ),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list(
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "**Table 2. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity**"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table2
```

## Table 3: combining tables 1 & 2 (using add function, will be the actual table 1 in the paper)

Rophence>>> so this is not the right way to combine tables 1 & 2, use the table 2 code above and all you need to add is the  add_overall() %>% function before the add_p() function

Kindy update your RMD and be sure to finish the rest of the tasks assigned yesterday and today!

Happy coding!

```{r}
# FIXED TABLE 3 GOES HERE
```

# Export as word document
```{r}
  library(flextable)
  table1 %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "table_1.docx")
  
  table2 %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "table_2.docx")
  
  combined_table %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "Output/Ria_table_3.docx")

```

## Ria's correlation table
```{r}
# Selecting all numerical data
  midus_corr <- midus_data %>% 
                select("income", "num_chron", "efficacy_STD",
                      "grimage2_STD", "dunedinpace_STD",
                      "cancerrisk_STD", "age", "bmi")

# Converting midus_corr to a matrix
  midus_matrix <- as.matrix(midus_corr)

# Calculating p-values and Pearson's Correlation Coefficient
  matrix <- Hmisc::rcorr(midus_matrix)
  
# Extract p-values and coefficients
  p <- matrix$P #p-values
  coeff <- matrix$r #correlation coefficients
  
# Adding starts to indicate significance in matrix
significance <- matrix("", nrow = nrow(p), ncol = ncol(p))
significance[p < 0.05]  <- "*"
significance[p < 0.01]  <- "**"
significance[p < 0.001] <- "***"
diag(significance) <- "" # Ensuring diagonal values have are empty

# Rounding correlation coefficients
coeff_rounded <- format(round(coeff, 3), nsmall = 3)

# Combining newly rounded values with stars
r_and_sig <- matrix(paste0(coeff_rounded, significance), ncol = ncol(coeff))
rownames(r_and_sig) <- rownames(coeff)
colnames(r_and_sig) <- colnames(coeff)

# Making copy of updated matrix (with starred formatting & rounding)
lower_values <- r_and_sig

# Replace the upper triangle with blank spaces
lower_values[upper.tri(lower_values, diag = FALSE)] <- ""

# View the final result
print(as.data.frame(lower_values))

```



## Ria's Imputation Code
# Getting started
```{r}
# Loading packages
library(tidyverse)    # Data wrangling and visualization
library(mice)         # Multiple imputation
library(missRanger)   # Random forest imputation
library(VIM)

# Set seed for reproducibility
set.seed(0864)

# Load dataset
data_dir  <- file.path(project_dir, "Data")
output_dir  <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
library(dplyr)
midus_data <- readRDS(analysis_path) #Canc_data: data from MIDUS cancer subproject
head(midus_data)
dim(midus_data)
```


# Subsetting data
```{r}
# Compiling relevant confounders in vector
# i.e. subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates
midus_vars <- midus_data %>% 
              select("income", "num_chron", "famcancer", "efficacy_STD",
                     "marital", "education", "grimage2_STD", "dunedinpace_STD",
                     "cancerrisk_STD", "race_eth", "age", "sex",
                     "smoking", "alcohol", "bmi") %>%
  # Categorical variables should be coded as factors in order to invoke polyreg
  mutate(
      race_eth = factor(race_eth),
      alcohol = factor(alcohol),
      sex = factor(sex),
      education = factor(education, ordered = TRUE),
      smoking = factor(smoking),
      marital = factor(marital),
      famcancer = factor(famcancer)
    )
              

```

# Summarizing missing data
```{r}
# Summarizing missing data
md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns 
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

# Missing data summary
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}
# missing_summary 
missing_summary(midus_data, names(midus_vars))

```

# Creating a predictor matrix
```{r}
# Creating variable to store matrix
matrix_vec <- make.predictorMatrix(midus_vars)

# Generating predictor matrix
pred_matrix <- mice(midus_vars, 
                    m = 20, 
                    maxit = 10, 
                    predictorMatrix = matrix_vec)

# Extracting an imputation to print predictor matrix
print_matrix <- pred_matrix$predictorMatrix

print_matrix
```

# Adjusting methods
```{r}
# Creating a methods object
midus_methods <- make.method(midus_vars) 
midus_methods # Looking to check method types for categorical vars

# Assigning new methods to categorical variables
midus_methods["education"] <- "polyreg" #polyreg used for categorical values 
# with 2+ categories
midus_methods["famcancer"] <- "logreg" #logreg for nominal categorical values
midus_methods["marital"] <- "polyreg"
midus_methods # Checking to ensure variables are updated
```

# Conducting the imputation using mice
```{r}

imp <- mice(
  midus_vars, # data: what we want to impute
  m = 20, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

```


```{r}
# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Combine the non-imputed variables with the imputed confounders
df_mice_imputed <- keep %>%
  bind_cols(df_mice_completed) #binding


missing_summary(df_mice_imputed, names(df_mice_imputed))
```

