---
title: "PQAR 2026 Research Project"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing necessary R packages
```{r}
# COMMENT ON THESE LIBRARIES
library("gtsummary")
library("flextable")
library("broom")
library("officer")
library("mice")
library("corrr")
library("tibble")
library("misty")
library("haven")
library("dplyr")
library("Hmisc")
library("modelsummary")
library("gt")
theme_gtsummary_journal(journal = "jama")
```


Classification of variables:
- Demographic information: age, sex, race/ethnicity, marital status, education, income
- Physical considerations: smoking, number of chronic diseases, BMI, alcohol, family cancer history
- Psychological considerations: efficacy
- Biological data: dunedinpace, grimage2
- Dependent variable of this study: cancer risk


# Reading in data and creating new dataframe based on demographic characteristics
```{r}
# Please comment your code so that someone else can follow along, for example, the code below is getting project working directory, also don't indent your code, add a line space to separate different lines of codes

# Get current project directory (RO)
project_dir <- getwd()

# Get data directory
data_dir  <- file.path(project_dir, "Data")
output_dir  <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
  
# loading in the datatset  
midus_data <- readRDS(analysis_path) #midus_data: data from MIDUS cancer subproject
  
# Adding descriptive labels to data frame >> you did not need to do this, as you are subsetting the dataset just to create labels, yet subseqent anlysis will require some of the standardized variables not included in this subset. One way to add labels is using the label function in tbl_summary. What you did here creates spaces in Variable names and variable names should not have spaces, so use label function within tbl_summary

```
 
**Before doing any further analysis below, remove the one participant with missing outcome. Also format the code for saving as word document. Your outputs are not aligned within the pagewidth, chnage to landscape, use autofit, e.t.c >> Rophence**


Create ‘Table 1' univariate statistics on the study population


```{r}
# Generate a basic summary table
theme_gtsummary_journal(journal = "jama")


table1 <- midus_data %>%
# Removing row with NA value for grimage2_STD
  filter(
    !is.na(grimage2_STD)
  ) %>%
# Mutating continuous variables to include "Missing" category
mutate(
    marital = as.character(marital),
    marital = if_else(is.na(marital), "Missing", marital),
    marital = factor(
      marital,
      levels = c("Married", "Divorced/Separated/Widowed", "Never married", "Missing")
    ),

    education = as.character(education),  # Fixed case to lowercase
    education = if_else(is.na(education), "Missing", education),
    education = factor(
      education,
      levels = c("High school or less", "Some college/associate's degree", "College degree or higher", "Missing")
    ),
    
    famcancer = as.character(famcancer), # Fixed name from "Family Cancer History"
    famcancer = if_else(is.na(famcancer), "Missing", famcancer),
    famcancer = factor(famcancer, levels = c("No", "Yes", "Missing"))
  ) %>%

  # Selecting for variables that will displayed in table (Don't include number of chronic conditions in all analysis >>> Rophence)
   select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%

  # Generating the table summary
  tbl_summary( 
    # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race and Ethnicity",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    # Assigning variable type (categorical or continuous)
    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer", "race_eth")) ~ "categorical",
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),
    
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
    
              digits = list(all_continuous() ~ 2,
                            all_categorical() ~ c(0, 1)), 
              missing = "ifany",
              missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
              missing_text = "Missing") %>%  # here you can now use  missing_text = "Missing" if you have coded categorical variables with missing values as Missing, previously it was missing = "ifany" which also shows missing values in continouous variables, and we do not want that.
  modify_caption("Table 1. Overall Descriptive Characteristics of MIDUS Participants") %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels() 

  table1

```


# Table 2: fixed

```{r}
# Table 2: Descriptive statistics by Race and Ethnicity
# No overall column.
# P-values compare each variable across Race/Ethnicity groups.
# Chi-square tests are used for categorical variables to avoid Fisher exact test errors.

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%
  
  tbl_summary(
        # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    by = race_eth,

    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer")) ~ "categorical",
                
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list(
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "Table 2. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table2
```


## Table 3: fixed

```{r}
# Table 3: Descriptive statistics by Race and Ethnicity
# Combines Tables 1 & 2
# Includes Overall column

table3 <- midus_data %>%
  select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%
  
  tbl_summary(
        # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    by = race_eth,

    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer")) ~ "categorical",
                
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_overall() %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list(
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "Table 3. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table3
```

# Export as word document
```{r}
  library(flextable)
  sect_properties <- prop_section(
    page_size = page_size(orient = "landscape"),
    page_margins = page_mar(top = 0.5, bottom = 0.5, left = 0.5, right = 0.5)
  )
  table1 %>%
    as_flex_table() %>%
    autofit() %>%
    flextable::save_as_docx(path = "Output/Ria_table_1_descriptive_characteristics.docx")
  
  table2 %>%
    as_flex_table() %>%
    autofit() %>%
    flextable::save_as_docx(path = "Output/Ria_table_2_stratified_descriptive_characteristics.docx")

  table3_fix <- table3 %>%
    as_flex_table() %>%
    # Forces fixed layout
    flextable::set_table_properties(layout = "fixed") %>%
  
    # Increases width of cells for text-wrapping
    flextable::width(j = "label", width = 2) %>%
    
    # Setting uniform width for other columns
    flextable::width(j = 2:flextable::ncol_keys(.), width = 0.8) %>%
  
    # Shrinking font and padding
    flextable::fontsize(size = 9, part = "all") %>%
    flextable::padding(padding.top = 0.5, padding.bottom = 0.5, part = "all") %>%
    
    flextable::save_as_docx(path = "Output/Ria_table_3_overall_stratified_descriptive_characteristics.docx")

```

## Ria's correlation table
```{r}
# Selecting all numerical data (Don't include chronic conditions in all analysis >> Rophence)
  midus_corr <- midus_data %>% 
                select("income", "num_chron", "efficacy_STD",
                      "grimage2_STD", "dunedinpace_STD",
                      "cancerrisk_STD", "age", "bmi")

# Converting midus_corr to a matrix
  midus_matrix <- as.matrix(midus_corr)

# Calculating p-values and Pearson's Correlation Coefficient
  matrix <- Hmisc::rcorr(midus_matrix)
  
# Extract p-values and coefficients
  p <- matrix$P #p-values
  coeff <- matrix$r #correlation coefficients
  
# Adding starts to indicate significance in matrix
significance <- matrix("", nrow = nrow(p), ncol = ncol(p))
significance[p < 0.05]  <- "*"
significance[p < 0.01]  <- "**"
significance[p < 0.001] <- "***"
diag(significance) <- "" # Ensuring diagonal values have are empty

# Rounding correlation coefficients
coeff_rounded <- format(round(coeff, 3), nsmall = 3)

# Combining newly rounded values with stars
r_and_sig <- matrix(paste0(coeff_rounded, significance), ncol = ncol(coeff))
rownames(r_and_sig) <- rownames(coeff)
colnames(r_and_sig) <- colnames(coeff)

# Making copy of updated matrix (with starred formatting & rounding)
lower_values <- r_and_sig

# Replace the upper triangle with blank spaces
lower_values[upper.tri(lower_values, diag = FALSE)] <- ""

# 2. Convert matrix to data frame safely (keeping column names raw)
df <- as.data.frame(lower_values)

# 3. Create the lookup vector for mapping labels
var_labels <- c(
  "income"          = "Annual Income ($)",
  "num_chron"       = "Number of Chronic Conditions",
  "efficacy_STD"    = "Self Efficacy",
  "grimage2_STD"    = "GrimAge2",
  "dunedinpace_STD" = "DunedinPACE",
  "cancerrisk_STD"  = "Perceived Cancer Risk",
  "age"             = "Age",
  "bmi"             = "Body Mass Index (kg/m²)"
)

# 4. Move row names into a brand-new column using the clean mapped labels
df <- cbind(
  Variable = var_labels[rownames(lower_values)], 
  df
)
rownames(df) <- NULL  # Clear old row names index attribute

# 5. Build and style the Flextable (JAMA theme)
jama_border <- fp_border(color = "black", width = 1)

jama_ft <- flextable(df) %>%
  set_header_labels(
    Variable        = "Variable",                  
    income          = "Annual Income ($)",
    num_chron       = "Number of Chronic Conditions",
    efficacy_STD    = "Self Efficacy",
    grimage2_STD    = "GrimAge2",
    dunedinpace_STD = "DunedinPACE",
    cancerrisk_STD  = "Perceived Cancer Risk",
    age             = "Age",
    bmi             = "Body Mass Index (kg/m²)"
  ) %>%
  border_remove() %>%
  hline_top(border = jama_border, part = "header") %>%
  hline_bottom(border = jama_border, part = "header") %>%
  hline_bottom(border = jama_border, part = "body") %>%
  fontsize(size = 9, part = "all") %>%
  font(fontname = "Arial", part = "all") %>%
  align(j = 1, align = "left", part = "all") %>%
  align(j = 2:ncol(df), align = "center", part = "all") %>%
  bold(part = "header") %>%
  bold(j = 1, part = "body")

# Setting jama-theme text style
jama_title_style <- fp_text(
  font.size = 11, 
  bold = TRUE, 
  font.family = "Arial"
)

jama_text_style <- fp_text(
  font.size = 11, 
  bold = FALSE, 
  font.family = "Arial"
)

# Generating word doc
doc <- read_docx() # Creates empty word document

doc <- doc %>%
  body_add_fpar(
    fpar(
      ftext("Table 2. ", prop = jama_title_style),
      ftext("Zero-order correlations among perceived cancer risk and epigenetic aging measures", prop = jama_text_style)
    )
  ) %>%
  body_add_flextable(jama_ft)

# Exporting final output as word document
print(doc, target = "Output/Ria_correlation_matrix.docx")

```



## Ria's Imputation Code
# Getting started
```{r}
# Loading packages
library(tidyverse)    # Data wrangling and visualization
library(mice)         # Multiple imputation
library(missRanger)   # Random forest imputation
library(VIM)
```

 I commented this out, as you don't need to load the data again (Roph)
 
```{r, eval=FALSE}
# Set seed for reproducibility
set.seed(0864)

# Load dataset
#data_dir  <- file.path(project_dir, "Data")
#output_dir  <- file.path(project_dir, "Output")
#analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
#library(dplyr)
#midus_data <- readRDS(analysis_path) #Canc_data: data from MIDUS cancer subproject
#head(midus_data)
#dim(midus_data)
```


# Subsetting data
```{r}
# Compiling relevant confounders in vector
# i.e. subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates

# checking for row containing NA value in grimAge2_STD
# print(is.na(midus_data$grimage2_STD))

midus_vars <- midus_data %>% 
              select("income", "famcancer", "efficacy",
                     "marital", "education", "grimage2_STD", "dunedinpace_STD",
                     "cancerrisk_STD", "race_eth", "age", "sex",
                     "smoking", "alcohol", "bmi") %>%
  # Removing row with NA value for grimage2_STD
  filter(
    !is.na(grimage2_STD)
  ) %>%
  # Categorical variables should be coded as factors in order to invoke polyreg
  mutate(
      race_eth = factor(race_eth),
      alcohol = factor(alcohol),
      sex = factor(sex),
      education = factor(education, ordered = TRUE),
      smoking = factor(smoking),
      marital = factor(marital),
      famcancer = factor(famcancer)
    )
              

# confirming NA value has been removed
# print(is.na(midus_vars$grimage2_STD))
```

# Summarizing missing data
```{r}
# Summarizing missing data
md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns 
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

# Missing data summary
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}
# missing_summary 
missing_summary(midus_data, names(midus_vars))

```

# Creating a predictor matrix
```{r}
# Creating variable to store matrix
matrix_vec <- make.predictorMatrix(midus_vars)

# Generating predictor matrix, I changed this to maxit = 0, and commented out m=20, as we are not running mice here but creating the predictor matrix (Rophence). You code was still fine even with these mistake
pred_matrix <- mice(midus_vars, 
                   # m = 20, 
                    maxit = 0, 
                    predictorMatrix = matrix_vec)

# Extracting an imputation to print predictor matrix
print_matrix <- pred_matrix$predictorMatrix

#print_matrix
```

# Adjusting methods
```{r}
# Creating a methods object
midus_methods <- make.method(midus_vars) 
midus_methods # Looking to check method types for categorical vars

# Assigning new methods to categorical variables
midus_methods["education"] <- "polr" # changed this from "polyreg" as education is an ordered categorical variable 
midus_methods["famcancer"] <- "logreg" #logreg for nominal categorical values
midus_methods["marital"] <- "polyreg" 
midus_methods # Checking to ensure variables are updated
```

# Conducting the imputation using mice
```{r}

imp <- mice(
  midus_vars, # data: what we want to impute
  m = 10, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

```


```{r}
# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Verifying that % missing is now zero
missing_summary(df_mice_completed, names(df_mice_completed))
```

## Ria's regression models


```{r}
# GrimAge2 & DunedinPACE Model 1
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + marital status
# + education + income
# Outcome: GrimAge2 & DunedinPACE Standardized Scores

# Calculating regression statistics for grimage2
m1_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital,
    data = df_mice_completed)

# Calculating regression statistics for dunedinPACE
m1_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital,
    data = df_mice_completed)


# GrimAge2 & DunedinPACE Models 2
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family cancer history
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m2_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income,
    data = df_mice_completed)
# Calculating regression statistics for DunedinPACE
m2_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income,
    data = df_mice_completed)



# GrimAge2 & DunedinPACE Model 3
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family history of cancer
# + self-efficacy for cancer prevention
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m3_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer,
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m3_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer,
    data = df_mice_completed)


#GrimAge2 & DunedinPACE Models 4
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family history of cancer
# + self-efficacy for cancer prevention + smoking + alcohol use
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m4_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer
    + efficacy,
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m4_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer
    + efficacy,
    data = df_mice_completed)


# GrimAge2 & DunedinPACE Models 5
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + self efficacy for cancer prevention
# + family history of cancer + smoking + alcohol + BMI
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m5_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD 
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi, 
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m5_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi, 
    data = df_mice_completed)


# GrimAge2 & DunedinPACE Models 6
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + self efficacy for cancer prevention
# + family history of cancer + smoking + alcohol + BMI + cancerrisk x race_eth
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m6_grimage2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth 
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi,
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m6_dunedinpace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi, 
    data = df_mice_completed) %>%

```
# Trying to include interaction terms within models 1-5
```{r}
# Calculating regression statistics for grimage2
m1_grimage_2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth 
    + age 
    + sex 
    + marital,
    data = df_mice_completed)

# Calculating regression statistics for dunedinPACE
m1_dunedin_pace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital,
    data = df_mice_completed)


# GrimAge2 & DunedinPACE Models 2
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family cancer history
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m2_grimage_2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income,
    data = df_mice_completed)
# Calculating regression statistics for DunedinPACE
m2_dunedin_pace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + race_eth
    + marital
    + education
    + income,
    data = df_mice_completed)



# GrimAge2 & DunedinPACE Model 3
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family history of cancer
# + self-efficacy for cancer prevention
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m3_grimage_2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer,
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m3_dunedin_pace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer,
    data = df_mice_completed)


#GrimAge2 & DunedinPACE Models 4
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + family history of cancer
# + self-efficacy for cancer prevention + smoking + alcohol use
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m4_grimage_2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy,
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m4_dunedin_pace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy,
    data = df_mice_completed)


# GrimAge2 & DunedinPACE Models 5
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + self efficacy for cancer prevention
# + family history of cancer + smoking + alcohol + BMI
# Outcome: GrimAge2 & DunedinPACE Standardized Scores
# Calculating regression statistics for GrimAge2
m5_grimage_2 <- lm(
    grimage2_STD ~ 
    cancerrisk_STD * race_eth 
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi, 
    data = df_mice_completed)

# Calculating regression statistics for DunedinPACE
m5_dunedin_pace <- lm(
    dunedinpace_STD ~ 
    cancerrisk_STD * race_eth
    + age 
    + sex 
    + marital
    + education
    + income
    + famcancer
    + efficacy
    + smoking
    + alcohol
    + bmi, 
    data = df_mice_completed)


# combining models for dunedin & grimage
dunedin_models_interact <- list(
  m1_dunedin_pace, m2_dunedin_pace, m3_dunedin_pace, 
  m4_dunedin_pace, m5_dunedin_pace#, m6_dunedinpace
)

grimage_models_interact <- list(
    m1_grimage_2, m2_grimage_2, m3_grimage_2,
```


```{r}
m4_grimage_2, m5_grimage_2#, m6_grimage2
)

# exporting models into doc

# Create dataframe of empty row with model 6 interaction term subheading
df_canc_interaction_1 <- data.frame(matrix(" ", nrow = 1, ncol = 6))
#rownames(df_canc_interaction) <- c("Perceived Cancer Risk * race_eth")
colnames(df_canc_interaction_1) <- c("part", "term", "model", "Est.", "({conf.low}, {conf.high})", "p")
df_canc_interaction_1$term <- "Perceived cancer risk (std.) * race_eth"

# Assigning model six subheader to specific row: 
attr(df_canc_interaction_1, "position") <- 7

# Grimage table in wide format
grimage_wide_1 <- modelsummary(
  models = grimage_models_interact, 
  output = "data.frame",
  fmt = 2,
  shape = term + model ~ statistic, # Makes Beta and P-value horizontal columns
  statistic = c("({conf.low}, {conf.high})", "p.value"),
  stars = TRUE,
  coef_map = c(
    "cancerrisk_STD" = "Perceived cancer risk (std.)",
    "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (std.) * non-Hispanic Black",
    "cancerrisk_STD:race_ethHispanic" = "Perceived cancer risk (std.) * Hispanic",
    "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (std.) * non-Hispanic Other"
    ),
  gof_omit = ".*",
  add_rows = df_canc_interaction_1
) %>% 
  select(-part) %>%
  rename(
    GrimAge2_Beta = Est., 
    `GrimAge2_CI` = `({conf.low}, {conf.high})`,
    `GrimAge2_p-value` = p,
  )


# DunedinPACE table in wide format
dunedin_wide_1 <- modelsummary(
  models = dunedin_models_interact, 
  output = "data.frame", 
  fmt = 2,
  shape = term + model ~ statistic, 
  statistic = c("({conf.low}, {conf.high})", "p.value"),
  stars = TRUE,
  coef_map = c(
    "cancerrisk_STD" = "Perceived cancer risk (std.)",
    "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (std.) * non-Hispanic Black",
    "cancerrisk_STD:race_ethHispanic" = "Perceived cancer risk (std.) * Hispanic",
    "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (std.) * non-Hispanic Other"
    ),
  gof_omit = ".*",
  add_rows = df_canc_interaction_1
) %>% 
  select(-part) %>%
  rename(
    DunedinPace_Beta = Est., 
    `DunedinPACE_CI` = `({conf.low}, {conf.high})`,
    `DunedinPace` = p,
  )


# Merging tables by the predictor name ("term") and model number
final_table_data_1 <- left_join(grimage_wide_1, dunedin_wide_1, by = c("term", "model")) %>%
  #Renaming model column
  mutate(model = dplyr::replace_values(
    model,
    "(1)" ~ "1",
    "(2)" ~ "2",
    "(3)" ~ "3",
    "(4)" ~ "4",
    "(5)" ~ "5",
    "(6)" ~ "6"
  ))


# 4. Build the Flextable
ft <- flextable(final_table_data_1) %>%
  set_caption(
    caption = "Table 2: Adjusted associations between perceived cancer risk 
              and epigenetic aging among MIDUS II sample"
  ) %>%
  # Adding group headers
  add_header_row(
    values = c("", "GrimAge2", "DunedinPACE"),
    colwidths = c(2, 3, 3)
  ) %>%
  set_header_labels(
    model = "Model",
    term = "Characteristic",
    GrimAge2_Beta = "Beta",
    `GrimAge2_CI` = "95% CI",
    `GrimAge2_p-value` = "p-value",
    DunedinPace_Beta = "Beta",
    `DunedinPACE_CI` = "95% CI",
    `DunedinPace` = "p-value"
  ) %>%
  add_footer_lines(
    values = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) %>%
  # italicizing legend
  italic(part = "footer")

# Initialize a blank Word Document
doc <- read_docx() %>%
  body_add_flextable(ft) 
  
# Export file
print(doc, target = "Output/ria_interaction_models.docx")
```



# Exporting models as table in word doc
```{r}
# Combining dunedin and grimage models into two separate vectors
dunedin_models <- list(
  m1_dunedinpace, m2_dunedinpace, m3_dunedinpace, 
  m4_dunedinpace, m5_dunedinpace#, m6_dunedinpace
)

grimage_models <- list(
  m1_grimage2, m2_grimage2, m3_grimage2, 
  m4_grimage2, m5_grimage2#, m6_grimage2
)

```


```{r}
# Create dataframe of empty row with model 6 interaction term subheading
df_canc_interaction <- data.frame(matrix(" ", nrow = 1, ncol = 6))
#rownames(df_canc_interaction) <- c("Perceived Cancer Risk * race_eth")
colnames(df_canc_interaction) <- c("part", "term", "model", "Est.", "({conf.low}, {conf.high})", "p")
df_canc_interaction$term <- "Perceived cancer risk (std.) * race_eth"

# Assigning model six subheader to specific row: 
attr(df_canc_interaction, "position") <- 7

# Grimage table in wide format
grimage_wide <- modelsummary(
  models = append(grimage_models, list(m6_grimage2)), 
  output = "data.frame",
  fmt = 2,
  shape = term + model ~ statistic, # Makes Beta and P-value horizontal columns
  statistic = c("({conf.low}, {conf.high})", "p.value"),
  stars = TRUE,
  coef_map = c(
    "cancerrisk_STD" = "Perceived cancer risk (std.)",
    "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (std.) * non-Hispanic Black",
    "cancerrisk_STD:race_ethHispanic" = "Perceived cancer risk (std.) * Hispanic",
    "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (std.) * non-Hispanic Other"
    ),
  gof_omit = ".*",
  add_rows = df_canc_interaction
) %>% 
  select(-part) %>%
  rename(
    GrimAge2_Beta = Est., 
    `GrimAge2_CI` = `({conf.low}, {conf.high})`,
    `GrimAge2_p-value` = p,
  )


# DunedinPACE table in wide format
dunedin_wide <- modelsummary(
  models = append(dunedin_models, list(m6_dunedinpace)), 
  output = "data.frame", 
  fmt = 2,
  shape = term + model ~ statistic, 
  statistic = c("({conf.low}, {conf.high})", "p.value"),
  stars = TRUE,
  coef_map = c(
    "cancerrisk_STD" = "Perceived cancer risk (std.)",
    "cancerrisk_STD:race_ethnon-Hispanic Black" = "Perceived cancer risk (std.) * non-Hispanic Black",
    "cancerrisk_STD:race_ethHispanic" = "Perceived cancer risk (std.) * Hispanic",
    "cancerrisk_STD:race_ethnon-Hispanic Other" = "Perceived cancer risk (std.) * non-Hispanic Other"
    ),
  gof_omit = ".*",
  add_rows = df_canc_interaction
) %>% 
  select(-part) %>%
  rename(
    DunedinPace_Beta = Est., 
    `DunedinPACE_CI` = `({conf.low}, {conf.high})`,
    `DunedinPace` = p,
  )


# Merging tables by the predictor name ("term") and model number
final_table_data <- left_join(grimage_wide, dunedin_wide, by = c("term", "model")) %>%
  #Renaming model column
  mutate(model = dplyr::replace_values(
    model,
    "(1)" ~ "1",
    "(2)" ~ "2",
    "(3)" ~ "3",
    "(4)" ~ "4",
    "(5)" ~ "5",
    "(6)" ~ "6"
  ))


# 4. Build the Flextable
ft <- flextable(final_table_data) %>%
  set_caption(
    caption = "Table 2: Adjusted associations between perceived cancer risk 
              and epigenetic aging among MIDUS II sample"
  ) %>%
  # Adding group headers
  add_header_row(
    values = c("", "GrimAge2", "DunedinPACE"),
    colwidths = c(2, 3, 3)
  ) %>%
  set_header_labels(
    model = "Model",
    term = "Characteristic",
    GrimAge2_Beta = "Beta",
    `GrimAge2_CI` = "95% CI",
    `GrimAge2_p-value` = "p-value",
    DunedinPace_Beta = "Beta",
    `DunedinPACE_CI` = "95% CI",
    `DunedinPace` = "p-value"
  ) %>%
  add_footer_lines(
    values = "* p < 0.05, ** p < 0.01, *** p < 0.001"
  ) %>%
  # italicizing legend
  italic(part = "footer")

# Initialize a blank Word Document
doc <- read_docx() %>%
  body_add_flextable(ft) 
  
# Export file
print(doc, target = "Output/ria_overall_models.docx")
```

# Grace code: adding stylistic details
```{r}
# =========================================================================
# 1. REGRESSION MODELS (Overall Sample)
# =========================================================================
df_overall <- df_mice_completed

# --- Outcome: GrimAge2 ---
grim_ov_m1 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital, data = df_overall)
grim_ov_m2 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income, data = df_overall)
grim_ov_m3 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer, data = df_overall)
grim_ov_m4 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy, data = df_overall)
grim_ov_m5 <- lm(grimage2_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)


# --- Outcome: DunedinPACE ---
pace_ov_m1 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital, data = df_overall)
pace_ov_m2 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income, data = df_overall)
pace_ov_m3 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer, data = df_overall)
pace_ov_m4 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy, data = df_overall)
pace_ov_m5 <- lm(dunedinpace_STD ~ cancerrisk_STD * race_eth + age + sex + marital + education + income + famcancer + efficacy + smoking + alcohol + bmi, data = df_overall)


# Create two flat lists containing all 6 models
grim_list <- list(grim_ov_m1, grim_ov_m2, grim_ov_m3, grim_ov_m4, grim_ov_m5)
pace_list <- list(pace_ov_m1, pace_ov_m2, pace_ov_m3, pace_ov_m4, pace_ov_m5)

# Variable mappings
# Variable mappings (Updated to handle both R naming conventions)
variables_to_show <- c(
  # Main Effect
  "cancerrisk_STD"                              = "Perceived Cancer Risk (Main Effect)",
  
  # Interaction terms (Continuous : Factor order)
  "cancerrisk_STD:race_ethnon-Hispanic Black"   = "Perceived cancer risk * non-Hispanic Black",
  "cancerrisk_STD:race_ethHispanic"             = "Perceived cancer risk * Hispanic",
  "cancerrisk_STD:race_ethnon-Hispanic Other"   = "Perceived cancer risk * non-Hispanic Other",
  
  # Interaction terms (Factor : Continuous order backup)
  "race_ethnon-Hispanic Black:cancerrisk_STD"   = "Perceived cancer risk * non-Hispanic Black",
  "race_ethHispanic:cancerrisk_STD"             = "Perceived cancer risk * Hispanic",
  "race_ethnon-Hispanic Other:cancerrisk_STD"   = "Perceived cancer risk * non-Hispanic Other"
)

# =========================================================================
# 2. EXTRACT NATIVE DATA FRAMES FROM MODELSUMMARY
# =========================================================================
df_grim <- modelsummary(
  grim_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}", statistic = "({conf.low}, {conf.high})",
  shape =  term ~ statistic,
  coef_map = variables_to_show
) %>%
  rename(
    beta1 = `(1) / Est.`,
    CI1 = `(1) / ({conf.low}, {conf.high})`,
    beta2 = `(2) / Est.`,
    CI2 = `(2) / ({conf.low}, {conf.high})`,
    beta3 = `(3) / Est.`,
    CI3 = `(3) / ({conf.low}, {conf.high})`,
    beta4 = `(4) / Est.`,
    CI4 = `(4) / ({conf.low}, {conf.high})`,
    beta5 = `(5) / Est.`,
    CI5 = `(5) / ({conf.low}, {conf.high})`
  ) %>%
  mutate(part = "GrimAge2")

df_pace <- modelsummary(
  pace_list, output = "data.frame", gof_omit = ".*", fmt = 2,
  estimate = "{estimate}{stars}", statistic = "({conf.low}, {conf.high})",
  shape =  term ~ statistic,
  coef_map = variables_to_show
) %>%
  rename(
    beta1 = `(1) / Est.`,
    CI1 = `(1) / ({conf.low}, {conf.high})`,
    beta2 = `(2) / Est.`,
    CI2 = `(2) / ({conf.low}, {conf.high})`,
    beta3 = `(3) / Est.`,
    CI3 = `(3) / ({conf.low}, {conf.high})`,
    beta4 = `(4) / Est.`,
    CI4 = `(4) / ({conf.low}, {conf.high})`,
    beta5 = `(5) / Est.`,
    CI5 = `(5) / ({conf.low}, {conf.high})`
  ) %>%
  mutate(part = "DunedinPACE")

# =========================================================================
# 3. MERGE AND CLEAN ROW DATA SHAPES
# =========================================================================
final_df <- bind_rows(df_grim, df_pace)

# =========================================================================
# 4. BUILD THE FINAL SEAMLESS FLEXTABLE
# =========================================================================
final_table <- flextable(final_df) %>%
  merge_v(j = "part") %>% 
  valign(j = "part", valign = "top") %>%
  # --- Bold Layout Modifications ---
  bold(part = "header") %>%                     # Bolds 'Outcome Metric', 'Predictor', and 'M1-M5'
  bold(j = "part", part = "body") %>%  # Bolds 'GrimAge2' and 'DunedinPACE' row labels
  add_header_row(
    values = c("", "", "Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
    colwidths = c(1, 1, 2, 2, 2, 2, 2)
  ) %>%
  set_header_labels(
    values = list(
    part = "Epigenetic Clock",
    term = "Characteristic",
    beta1 = "Beta",
    CI1 = "95% CI",
    beta2 = "Beta",
    CI2 = "95% CI",
    beta3 = "Beta",
    CI3 = "95% CI",
    beta4 = "Beta",
    CI4 = "95% CI",
    beta5 = "Beta",
    CI5 = "95% CI"
    )
  ) %>%
  add_footer_lines("CI = Confidence Interval\n+ p < 0.10, * p < 0.05, ** p < 0.01, *** p < 0.001.") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  set_caption(
    caption = "Table 4: Overall models with standardized beta coefficients (and 95% confidence intervals) from a series of linear regression models quantifying the association between perceived cancer risk and epigenetic aging among MIDUS II sample"
  ) 

```

