---
title: "PQAR 2026 Research Project"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing necessary R packages
```{r}
# COMMENT ON THESE LIBRARIES
library("gtsummary")
library("flextable")
library("broom")
library("officer")
library("mice")
library("corrr")
library("tibble")
library("misty")
library("haven")
library("dplyr")
library("Hmisc")
```


Classification of variables:
- Demographic information: age, sex, race/ethnicity, marital status, education, income
- Physical considerations: smoking, number of chronic diseases, BMI, alcohol, family cancer history
- Psychological considerations: efficacy
- Biological data: dunedinpace, grimage2
- Dependent variable of this study: cancer risk


# Reading in data and creating new dataframe based on demographic characteristics
```{r}
# Please comment your code so that someone else can follow along, for example, the code below is getting project working directory, also don't indent your code, add a line space to separate different lines of codes

# Get current project directory (RO)
project_dir <- getwd()

# Get data directory
data_dir  <- file.path(project_dir, "Data")
output_dir  <- file.path(project_dir, "Output")
analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
  
# loading in the datatset  
midus_data <- readRDS(analysis_path) #midus_data: data from MIDUS cancer subproject
  
# Adding descriptive labels to data frame >> you did not need to do this, as you are subsetting the dataset just to create labels, yet subseqent anlysis will require some of the standardized variables not included in this subset. One way to add labels is using the label function in tbl_summary. What you did here creates spaces in Variable names and variable names should not have spaces, so use label function within tbl_summary

```



Create ‘Table 1' univariate statistics on the study population


```{r}
# Generate a basic summary table
theme_gtsummary_journal(journal = "jama")


table1 <- midus_data %>%
# Mutating continuous variables to include "Missing" category
mutate(
    marital = as.character(marital),
    marital = if_else(is.na(marital), "Missing", marital),
    marital = factor(
      marital,
      levels = c("Married", "Divorced/Separated/Widowed", "Never married", "Missing")
    ),

    education = as.character(education),  # Fixed case to lowercase
    education = if_else(is.na(education), "Missing", education),
    education = factor(
      education,
      levels = c("High school or less", "Some college/associate's degree", "College degree or higher", "Missing")
    ),
    
    famcancer = as.character(famcancer), # Fixed name from "Family Cancer History"
    famcancer = if_else(is.na(famcancer), "Missing", famcancer),
    famcancer = factor(famcancer, levels = c("No", "Yes", "Missing"))
  ) %>%

  # Selecting for variables that will displayed in table
   select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%

  # Generating the table summary
  tbl_summary( 
    # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race and Ethnicity",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    # Assigning variable type (categorical or continuous)
    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer", "race_eth")) ~ "categorical",
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),
    
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
    
              digits = list(all_continuous() ~ 2,
                            all_categorical() ~ c(0, 1)), 
              missing = "ifany",
              missing_stat = "{N_miss} ({p_miss}%)",  # this adds the frequency and % missing
              missing_text = "Missing") %>%  # here you can now use  missing_text = "Missing" if you have coded categorical variables with missing values as Missing, previously it was missing = "ifany" which also shows missing values in continouous variables, and we do not want that.
  modify_caption("**Table 1. Overall Descriptive Characteristics of MIDUS Participants**") %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

  table1

```


# Table 2: fixed

```{r}
# Table 2: Descriptive statistics by Race and Ethnicity
# No overall column.
# P-values compare each variable across Race/Ethnicity groups.
# Chi-square tests are used for categorical variables to avoid Fisher exact test errors.

table2 <- midus_data %>%
  select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%
  
  tbl_summary(
        # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    by = race_eth,

    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer")) ~ "categorical",
                
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list(
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "**Table 2. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity**"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table2
```


## Table 3: fixed

```{r}
# Table 3: Descriptive statistics by Race and Ethnicity
# Combines Tables 1 & 2
# Includes Overall column

table3 <- midus_data %>%
  select(age, sex, race_eth, marital, education,
         smoking, alcohol, income, famcancer, num_chron, bmi,
         cancerrisk_STD, efficacy_STD, dunedinpace_STD, grimage2_STD
    ) %>%
  
  tbl_summary(
        # Assigning proper labels to variable names
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      marital ~ "Marital Status",
      education ~ "Education",
      smoking ~ "Smoking Status",
      alcohol ~ "Alcohol Consumption",
      income ~ "Income ($)",
      famcancer ~ "Family History of Cancer",
      num_chron ~ "Number of Chronic Conditions",
      bmi ~ "Body Mass Index (kg/m^2)",
      cancerrisk_STD ~ "Perceived Cancer Risk",
      efficacy_STD ~ "Self Efficacy",
      dunedinpace_STD ~ "DunedinPACE",
      grimage2_STD ~ "GrimAge2"
      ),
    
    by = race_eth,

    type = list(all_of(c("sex", "marital", "education", "smoking", 
                "alcohol", "famcancer")) ~ "categorical",
                
                all_of(c("age", "income", "bmi", "grimage2_STD", 
                "dunedinpace_STD", "num_chron",
                "cancerrisk_STD", "efficacy_STD")) ~ "continuous"),

    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),

    missing = "ifany",
    missing_text = "Missing",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_overall() %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_continuous() ~ "kruskal.test"
    ),
    test.args = list(
      all_tests("chisq.test") ~ list(simulate.p.value = TRUE, B = 2000)
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  modify_caption(
    "**Table 3. Descriptive Characteristics of MIDUS Participants by Race and Ethnicity**"
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table3
```

# Export as word document
```{r}
  library(flextable)
  table1 %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "Output/Ria_table_1_descriptive_characteristics.docx")
  
  table2 %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "Output/Ria_ttable_2_stratified_descriptive_characteristics.docx")
  
  table3 %>%
    as_flex_table() %>%
    flextable::save_as_docx(path = "Output/Ria_table_3_overall_stratified_descriptive_characteristics.docx")

```

## Ria's correlation table
```{r}
# Selecting all numerical data
  midus_corr <- midus_data %>% 
                select("income", "num_chron", "efficacy_STD",
                      "grimage2_STD", "dunedinpace_STD",
                      "cancerrisk_STD", "age", "bmi")

# Converting midus_corr to a matrix
  midus_matrix <- as.matrix(midus_corr)

# Calculating p-values and Pearson's Correlation Coefficient
  matrix <- Hmisc::rcorr(midus_matrix)
  
# Extract p-values and coefficients
  p <- matrix$P #p-values
  coeff <- matrix$r #correlation coefficients
  
# Adding starts to indicate significance in matrix
significance <- matrix("", nrow = nrow(p), ncol = ncol(p))
significance[p < 0.05]  <- "*"
significance[p < 0.01]  <- "**"
significance[p < 0.001] <- "***"
diag(significance) <- "" # Ensuring diagonal values have are empty

# Rounding correlation coefficients
coeff_rounded <- format(round(coeff, 3), nsmall = 3)

# Combining newly rounded values with stars
r_and_sig <- matrix(paste0(coeff_rounded, significance), ncol = ncol(coeff))
rownames(r_and_sig) <- rownames(coeff)
colnames(r_and_sig) <- colnames(coeff)

# Making copy of updated matrix (with starred formatting & rounding)
lower_values <- r_and_sig

# Replace the upper triangle with blank spaces
lower_values[upper.tri(lower_values, diag = FALSE)] <- ""

# 2. Convert matrix to data frame safely (keeping column names raw)
df <- as.data.frame(lower_values)

# 3. Create the lookup vector for mapping labels
var_labels <- c(
  "income"          = "Annual Income ($)",
  "num_chron"       = "Number of Chronic Conditions",
  "efficacy_STD"    = "Self Efficacy",
  "grimage2_STD"    = "GrimAge2",
  "dunedinpace_STD" = "DunedinPACE",
  "cancerrisk_STD"  = "Perceived Cancer Risk",
  "age"             = "Age",
  "bmi"             = "Body Mass Index (kg/m²)"
)

# 4. Move row names into a brand-new column using the clean mapped labels
df <- cbind(
  Variable = var_labels[rownames(lower_values)], 
  df
)
rownames(df) <- NULL  # Clear old row names index attribute

# 5. Build and style the Flextable (JAMA theme)
jama_border <- fp_border(color = "black", width = 1)

jama_ft <- flextable(df) %>%
  set_header_labels(
    Variable        = "Variable",                  
    income          = "Annual Income ($)",
    num_chron       = "Number of Chronic Conditions",
    efficacy_STD    = "Self Efficacy",
    grimage2_STD    = "GrimAge2",
    dunedinpace_STD = "DunedinPACE",
    cancerrisk_STD  = "Perceived Cancer Risk",
    age             = "Age",
    bmi             = "Body Mass Index (kg/m²)"
  ) %>%
  border_remove() %>%
  hline_top(border = jama_border, part = "header") %>%
  hline_bottom(border = jama_border, part = "header") %>%
  hline_bottom(border = jama_border, part = "body") %>%
  fontsize(size = 9, part = "all") %>%
  font(fontname = "Arial", part = "all") %>%
  align(j = 1, align = "left", part = "all") %>%
  align(j = 2:ncol(df), align = "center", part = "all") %>%
  bold(part = "header") %>%
  bold(j = 1, part = "body")

# Setting jama-theme text style
jama_title_style <- fp_text(
  font.size = 11, 
  bold = TRUE, 
  font.family = "Arial"
)

jama_text_style <- fp_text(
  font.size = 11, 
  bold = FALSE, 
  font.family = "Arial"
)

# Generating word doc
doc <- read_docx() # Creates empty word document

doc <- doc %>%
  body_add_fpar(
    fpar(
      ftext("Table 2. ", prop = jama_title_style),
      ftext("Zero-order correlations among perceived cancer risk and epigenetic aging measures", prop = jama_text_style)
    )
  ) %>%
  body_add_flextable(jama_ft)

# Exporting final output as word document
print(doc, target = "Output/Ria_correlation_matrix.docx")

```



## Ria's Imputation Code
# Getting started
```{r}
# Loading packages
library(tidyverse)    # Data wrangling and visualization
library(mice)         # Multiple imputation
library(missRanger)   # Random forest imputation
library(VIM)
```

 I commented this out, as you don't need to load the data again (Roph)
 
```{r, eval=FALSE}
# Set seed for reproducibility
set.seed(0864)

# Load dataset
#data_dir  <- file.path(project_dir, "Data")
#output_dir  <- file.path(project_dir, "Output")
#analysis_path <- file.path(data_dir, "midus_cancer_aging_analysis_dataset.rds")
#library(dplyr)
#midus_data <- readRDS(analysis_path) #Canc_data: data from MIDUS cancer subproject
#head(midus_data)
#dim(midus_data)
```


# Subsetting data
```{r}
# Compiling relevant confounders in vector
# i.e. subsetting the data to include the two standardized outcomes, 
# the standardized exposure, the moderator and the 11 covariates
midus_vars <- midus_data %>% 
              select("income", "num_chron", "famcancer", "efficacy_STD",
                     "marital", "education", "grimage2_STD", "dunedinpace_STD",
                     "cancerrisk_STD", "race_eth", "age", "sex",
                     "smoking", "alcohol", "bmi") %>%
  # Categorical variables should be coded as factors in order to invoke polyreg
  mutate(
      race_eth = factor(race_eth),
      alcohol = factor(alcohol),
      sex = factor(sex),
      education = factor(education, ordered = TRUE),
      smoking = factor(smoking),
      marital = factor(marital),
      famcancer = factor(famcancer)
    )
              

```

# Summarizing missing data
```{r}
# Summarizing missing data
md.pattern(midus_data) # Counting number of NA
# Pattern summary: 1087 rows of 1156 that are complete. 

md.pairs(midus_data) # Determining four missing patterns 
# Looking at mr / rm values helps understand which values have overlapping data
# to make them eligible for calculating imputations. 

# Missing data summary
missing_summary <- function(data, vars) {
  data %>%
    summarise(
      across(
        all_of(vars), #across all the given vars
        ~ mean(is.na(.x)) * 100 # percentage of variables w/ missing values
      )
    ) %>% #create a data set w/ names - variable names, values - percent missing
    pivot_longer(
      cols = everything(),
      names_to = "variable",
      values_to = "percent_missing"
    )
}
# missing_summary 
missing_summary(midus_data, names(midus_vars))

```

# Creating a predictor matrix
```{r}
# Creating variable to store matrix
matrix_vec <- make.predictorMatrix(midus_vars)

# Generating predictor matrix, I changed this to maxit = 0, and commented out m=20, as we are not running mice here but creating the predictor matrix (Rophence). You code was still fine even with these mistake
pred_matrix <- mice(midus_vars, 
                   # m = 20, 
                    maxit = 0, 
                    predictorMatrix = matrix_vec)

# Extracting an imputation to print predictor matrix
print_matrix <- pred_matrix$predictorMatrix

#print_matrix
```

# Adjusting methods
```{r}
# Creating a methods object
midus_methods <- make.method(midus_vars) 
midus_methods # Looking to check method types for categorical vars

# Assigning new methods to categorical variables
midus_methods["education"] <- "polr" # changed this from "polyreg" as education is an ordered categorical variable 
midus_methods["famcancer"] <- "logreg" #logreg for nominal categorical values
midus_methods["marital"] <- "polyreg" 
midus_methods # Checking to ensure variables are updated
```

# Conducting the imputation using mice
```{r}

imp <- mice(
  midus_vars, # data: what we want to impute
  m = 10, # number of imputations to perform
  seed = 0864,
  printFlag = FALSE,
  method = midus_methods
)

```


```{r}
# Extract the first completed imputed dataset
df_mice_completed <- complete(imp, action = 1)
df_mice_completed

# Verifying that % missing is now zero
missing_summary(df_mice_completed, names(df_mice_completed))
```

## Ria's regression models


```{r}

# Models for Hispanic participants
#GrimAge2 Model 1
# Covariates: Perceived cancer risk + age + sex + marital status
# Outcome: GrimAge2 Standardized Scores

# Subsetting data as data frames by race/ethnicity
hispanic_pop <- df_mice_completed[df_mice_completed$race_eth == "Hispanic", ]
# Calculating regression statistics
model1_hisp <- lm(grimage2_STD ~ cancerrisk_STD + age + sex + marital, data = hispanic_pop)
summary(model1_hisp)
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.70987 -0.17619 -0.08119  0.22925  0.83587 
# 
# Coefficients:
#                                    Estimate Std. Error t value Pr(>|t|)    
# (Intercept)                       -3.986771   0.262499 -15.188   <2e-16 ***
# cancerrisk_STD                     0.129533   0.072431   1.788   0.0824 .  
# age                                0.078912   0.004905  16.087   <2e-16 ***
# sexFemale                         -0.086788   0.118451  -0.733   0.4686    
# maritalDivorced/Separated/Widowed  0.334796   0.137500   2.435   0.0201 *  
# maritalNever married               0.146090   0.164090   0.890   0.3794    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.3553 on 35 degrees of freedom
# Multiple R-squared:  0.9093,	Adjusted R-squared:  0.8963 
# F-statistic: 70.18 on 5 and 35 DF,  p-value: < 2.2e-16

# Creating a plot for model 1
library(ggeffects)
ggpredict(model1_hisp, terms = c("age", "sex", "marital")) %>%
  plot(colors = "bw")

#GrimAge2 Model 2
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income
# Outcome: GrimAge2 Standardized Scores
# Calculating regression statistics
model2_hisp <- lm(grimage2_STD ~ 
                  cancerrisk_STD + age + sex + marital + education + income, 
                  data = hispanic_pop)
summary(model2_hisp)
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.68461 -0.19616 -0.01902  0.21238  0.78185 
# 
# Coefficients:
#                                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)                       -3.955e+00  3.224e-01 -12.266 1.23e-13 ***
# cancerrisk_STD                     9.297e-02  7.499e-02   1.240   0.2241    
# age                                7.929e-02  5.289e-03  14.991 5.06e-16 ***
# sexFemale                         -1.132e-01  1.189e-01  -0.952   0.3482    
# maritalDivorced/Separated/Widowed  3.061e-01  1.451e-01   2.110   0.0428 *  
# maritalNever married               1.921e-01  1.915e-01   1.003   0.3234    
# education.L                       -1.852e-01  1.023e-01  -1.811   0.0796 .  
# education.Q                        2.531e-02  1.152e-01   0.220   0.8275    
# income                            -2.908e-07  9.320e-07  -0.312   0.7571    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.3531 on 32 degrees of freedom
# Multiple R-squared:  0.9181,	Adjusted R-squared:  0.8976 
# F-statistic: 44.83 on 8 and 32 DF,  p-value: 3.133e-15


#GrimAge2 Model 3
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income 
# + number of chronic conditions + family history of cancer
# Outcome: GrimAge2 Standardized Scores
# Calculating regression statistics
model3_hisp <- lm(grimage2_STD ~ 
                  cancerrisk_STD + age + sex + marital + education + income 
                  + num_chron + famcancer, data = hispanic_pop)
summary(model3_hisp)
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.71593 -0.17367  0.00047  0.19394  0.78110 
# 
# Coefficients:
#                                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)                       -4.061e+00  3.408e-01 -11.916 6.65e-13 ***
# cancerrisk_STD                     8.897e-02  7.689e-02   1.157    0.256    
# age                                8.168e-02  5.772e-03  14.151 8.17e-15 ***
# sexFemale                         -1.217e-01  1.205e-01  -1.010    0.320    
# maritalDivorced/Separated/Widowed  2.517e-01  1.566e-01   1.608    0.118    
# maritalNever married               1.804e-01  1.982e-01   0.910    0.370    
# education.L                       -1.672e-01  1.052e-01  -1.590    0.122    
# education.Q                        1.615e-02  1.202e-01   0.134    0.894    
# income                            -3.862e-07  9.461e-07  -0.408    0.686    
# num_chron                          2.276e-02  2.304e-02   0.988    0.331    
# famcancerYes                      -1.091e-01  1.363e-01  -0.801    0.430    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.357 on 30 degrees of freedom
# Multiple R-squared:  0.9215,	Adjusted R-squared:  0.8953 
# F-statistic: 35.22 on 10 and 30 DF,  p-value: 7.557e-14

#GrimAge2 Model 4
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + self efficacy
# + number of chronic conditions + family history of cancer
# Outcome: GrimAge2 Standardized Scores
# Calculating regression statistics
model4_hisp <- lm(grimage2_STD ~ 
                  cancerrisk_STD + age + sex + marital + education + income 
                  + num_chron + famcancer + efficacy_STD, data = hispanic_pop)
summary(model4_hisp)
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.70954 -0.17250 -0.00084  0.19890  0.78793 
# 
# Coefficients:
#                                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)                       -4.054e+00  3.536e-01 -11.466 2.72e-12 ***
# cancerrisk_STD                     8.926e-02  7.826e-02   1.141    0.263    
# age                                8.159e-02  5.949e-03  13.715 3.31e-14 ***
# sexFemale                         -1.215e-01  1.225e-01  -0.992    0.330    
# maritalDivorced/Separated/Widowed  2.502e-01  1.600e-01   1.563    0.129    
# maritalNever married               1.782e-01  2.029e-01   0.878    0.387    
# education.L                       -1.681e-01  1.074e-01  -1.566    0.128    
# education.Q                        1.764e-02  1.233e-01   0.143    0.887    
# income                            -4.063e-07  9.858e-07  -0.412    0.683    
# num_chron                          2.285e-02  2.345e-02   0.974    0.338    
# famcancerYes                      -1.100e-01  1.389e-01  -0.792    0.435    
# efficacy_STD                      -6.882e-03  7.327e-02  -0.094    0.926    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.363 on 29 degrees of freedom
# Multiple R-squared:  0.9215,	Adjusted R-squared:  0.8918 
# F-statistic: 30.96 on 11 and 29 DF,  p-value: 4.633e-13

#GrimAge2 Model 5
# Covariates: Perceived cancer risk + age + sex + race/ethnicity + 
# marital status + education + income + self efficacy
# + number of chronic conditions + family history of cancer
# + smoking + alcohol + BMI
# Outcome: GrimAge2 Standardized Scores
# Calculating regression statistics
model5_hisp <- lm(grimage2_STD ~ 
                  cancerrisk_STD + age + sex + marital + education + income 
                  + num_chron + famcancer + efficacy_STD
                  + smoking + alcohol + bmi, data = hispanic_pop)
summary(model5_hisp)
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.50558 -0.10608 -0.00401  0.12109  0.56897 
# 
# Coefficients:
#                                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)                       -4.378e+00  5.812e-01  -7.532 1.19e-07 ***
# cancerrisk_STD                     1.627e-02  7.642e-02   0.213   0.8333    
# age                                7.897e-02  5.808e-03  13.597 1.76e-12 ***
# sexFemale                         -1.789e-02  1.281e-01  -0.140   0.8901    
# maritalDivorced/Separated/Widowed -3.917e-02  1.829e-01  -0.214   0.8323    
# maritalNever married               1.385e-01  1.831e-01   0.757   0.4570    
# education.L                       -5.742e-02  1.023e-01  -0.561   0.5801    
# education.Q                        6.176e-03  1.187e-01   0.052   0.9590    
# income                            -7.820e-07  9.024e-07  -0.867   0.3951    
# num_chron                          3.917e-03  2.212e-02   0.177   0.8610    
# famcancerYes                      -1.736e-01  1.380e-01  -1.258   0.2209    
# efficacy_STD                      -2.995e-02  7.563e-02  -0.396   0.6958    
# smokingPast                        2.368e-01  1.482e-01   1.598   0.1237    
# smokingCurrent                     7.842e-01  3.190e-01   2.458   0.0219 *  
# alcohol<1 day a week              -2.972e-02  1.667e-01  -0.178   0.8601    
# alcohol1-2 days a week            -2.895e-03  1.918e-01  -0.015   0.9881    
# alcohol3+ days a week              2.625e-01  1.927e-01   1.362   0.1863    
# bmi                                1.398e-02  1.604e-02   0.871   0.3926    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.3127 on 23 degrees of freedom
# Multiple R-squared:  0.9538,	Adjusted R-squared:  0.9197 
# F-statistic: 27.95 on 17 and 23 DF,  p-value: 2e-11

```

#